---
title: "Correlation Exploration: Heatmaps, scatter plots and other fun things"
author: "Florencia Zúñiga"
date: "7/10/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



####  1.  LOADING DATA                                                                                                      

## Loading packages

```{r}
library(readr)
library(rstudioapi)
library(BBmisc)

#Packages needed for the heatmap
library(pheatmap)
library("DESeq")
library(dendextend)

#Packages for the dendogram
library(ggplot2)
library(circlize)

#Packages for the correlogram
library(corrgram)
```



```{r,include = FALSE}

wd = ("/GitHub/project-02-group-05")
```


Reading the data
```{r}
Untreated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_untreated.rds"))
Treated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_treated.rds"))

Metadata = read.table(paste0(wd,"/data/NCI_TPW_metadata.tsv"), 
                      header = TRUE, sep ="\t", stringsAsFactors = TRUE)

Sensitivity <- readRDS(paste0(wd,"/data/NegLogGI50.rds"))


Basal <- readRDS(paste0(wd,"/data/CCLE_basalexpression.rds"))
Copynumber <- readRDS(paste0(wd,"/data/CCLE_copynumber.rds"))
Mutations <- readRDS(paste0(wd,"/data/CCLE_mutations.rds"))

Cellline_annotation = read.table(paste0(wd,"/data/cellline_annotation.tsv"), 
                                 header = TRUE, sep ="\t", stringsAsFactors = TRUE)
Drug_annotation = read.table(paste0(wd,"/data/drug_annotation.tsv"), 
                             header = TRUE, sep ="\t", stringsAsFactors = TRUE)
```

Transforming the data
```{r}
Treated <- as.data.frame(Treated)
Untreated <- as.data.frame(Untreated)
Sensitivity<- as.data.frame(Sensitivity)
```

Data normalization
```{r}
Untreated_norm <- apply(Untreated, 2, function(x){
  (x - mean(x)) / sd(x)
 })


Treated_norm <- apply(Treated, 2, function(x){
  (x - mean(x)) / sd(x)
 })


FC <- Treated - Untreated
FC_norm <- apply(FC, 2, function(x){
  (x - mean(x)) / sd(x)
 })
```

###   1.1 Biomarkers                                                                                                        
##  (1)  Creating Vorinostat
```{r}
#Untreated matrix
UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))

#Same with treated matrix
TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treated))

#Define Vorinostat-data: 
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]

#fold change matrix 
FC <- TreatedVorinostat - UntreatedVorinostat

#Sensitivity 
vorinostat_Sensitivity_alleZeilen= grep ('vorinostat', rownames(Sensitivity))
vorinostat_Sensitivity_data= Sensitivity[vorinostat_Sensitivity_alleZeilen,]
```

##  (2)  Creating FC Data -  Finding the Biomarkers

```{r}
FC <- TreatedVorinostat - UntreatedVorinostat

#We work with mean of the rows because we only want to compare the genes 
FC_meanrow= rowMeans(FC)

## Sorting the data 
#We work with absolute value to find the highest values, 
#because we want to have the most up and down regulated genes.
FC_abs= abs(FC_meanrow)

#We sort the values to get the 100 largest values 
sortedFC_abs <- sort(FC_abs, decreasing = TRUE)
sortedFC_abs <- as.matrix(sortedFC_abs)

#We select the first n for biomarkers 
biomarkers_FC30 = sortedFC_abs[1:30,]
biomarkers_FC30 <- as.matrix(biomarkers_FC30)

biomarkers_FC100 = sortedFC_abs[1:100,]
biomarkers_FC100 <- as.matrix(biomarkers_FC100)

## Creating a matrix with FC values, that are both positive and negative 
FC_both= cbind(FC_meanrow,FC_abs)
FC_both=as.data.frame(FC_both)

#Ordering this matrix 
FC_both_sorted <- FC_both[order(FC_both$FC_abs, decreasing = TRUE),]

#FC values of biomarkers: We select the first 100 of the sorted matrix. 
biomarkers_FC_values30 = FC_both_sorted[1:30,]

biomarkers_FC_values100 = FC_both_sorted[1:100,]

#Removing the absolute values
biomarkers_FC_values30 <- subset( biomarkers_FC_values30, select = -FC_abs)
biomarkers_FC_values30 = as.matrix(biomarkers_FC_values30)

biomarkers_FC_values100 <- subset( biomarkers_FC_values100, select = -FC_abs)
biomarkers_FC_values100 = as.matrix(biomarkers_FC_values100)

```

### 2.    Biomarkers found using the mean vs tissue                                          

##  2.1   Table with Biomarkers, difference beatween treated and untreated, and cell lines                    

## CREATING THE TABLE FOR 30 BIOMARKERS
```{r}
cor1_tab = FC [ which(row.names(FC) %in% rownames(biomarkers_FC_values30)), ]
```


Cleaning the names of the columns and rows:
COLUMNS ONLY WITH CELL LINE + CHANGE NAME OF GENE HIST, CHANGE CELL LINES NAME'S TO TISSUES

```{r}
col_names_cor1 = as.data.frame(strsplit(x=colnames(cor1_tab),split="_vorinostat"))
colnames (cor1_tab) = as.data.frame (t(col_names_cor1[1,]))[,1]

row_names_cor1 = as.data.frame(strsplit(x=rownames(cor1_tab),split="///HIST"))
rownames (cor1_tab) = as.data.frame (t(row_names_cor1[1,]))[,1]

cor1_tab = as.matrix(cor1_tab)
```

This table can only be read if it is a numerical matrix, so first we check if this is true

```{r}
class(cor1_tab)
```

```{r}
is.numeric(cor1_tab)
```

Now that we have the table, and we know that it is a numeric matrix, we can proceed to use it to use for the heatmap

## CREATING THE TABLE FOR 100 BIOMARKERS

```{r}
cor1.2_tab = FC [ which(row.names(FC) %in% rownames(biomarkers_FC_values100)), ]
```

Cleaning the names of the columns and rows:
COLUMNS ONLY WITH CELL LINE + CHANGE NAME OF GENE HIST, CHANGE CELL LINES NAME'S TO TISSUES

```{r}
col_names_cor1.2    = as.data.frame(strsplit(x=colnames(cor1.2_tab),split="_vorinostat"))
colnames (cor1.2_tab) = as.data.frame (t(col_names_cor1.2[1,]))[,1]

row_names_cor1.2    = as.data.frame(strsplit(x=rownames(cor1.2_tab),split="///HIST"))
rownames (cor1.2_tab) = as.data.frame (t(row_names_cor1.2[1,]))[,1]

cor1.2_tab = as.matrix(cor1.2_tab)
```

This table can only be read if it is a numerical matrix, so first we check if this is true

```{r}
class(cor1.2_tab)
```

```{r}
is.numeric(cor1.2_tab)
```

Now that we have the table, and we know that it is a numeric matrix, we can proceed to use it to use for the heatmap


#   2.2.1 Heatmap with 30 Biomarkers and 59 cell lines                                                        

This is how the heatmap looks making any modifications.
```{r}
pheatmap(cor1_tab)
```


# Scaling the rows for standard normal deviation

```{r, include=FALSE}
cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}

cor1_tab_norm <- t(apply(cor1_tab, 1, cal_z_score))
pheatmap(cor1_tab_norm)
colnames(cor1_tab_norm) == Metadata_V_after$cell
```


We can visualize a pheatmap with scaled rows using the standard normal distribution. Here we use tissue names instead of cell line names.
```{r}
cor1_tab_norm_tissue = cor1_tab_norm
colnames(cor1_tab_norm_tissue) = Metadata_V_after$tissue
pheatmap(cor1_tab_norm_tissue)
```
When we apply the function for standard normal distribution, the data changes its appearence in a way in which not many conclusions can be drawn. It is for this reason that, for the rest of the analysis, the biomarkers will not be normalized using the function of the code above. 

# Ploting a dendogram and cutting the tree
```{r}
cor1_hclust <- hclust(dist(cor1_tab), method = "complete")

as.dendrogram(cor1_hclust) %>%
  plot(horiz = TRUE)

cor1_hclust_tree <- cutree(tree = as.dendrogram(cor1_hclust), k = 2)
```

To which cluster do some of our genes belong?
```{r}
head(cor1_hclust_tree)
```

We can see that different genes belong to different clusters.


# Creating a dataframe, needed for annotations, and adding row annotations


Tissue Annotation: Creating a table with the tissue that corresponds to each cell line

```{r}
tissue <- Metadata[,-c(2:5)] 

row.names(tissue) <- tissue$sample
tissue <- tissue[-c(1)]


tissue = as.data.frame(tissue)
cor1_tissue <- tissue[-c(1:1579), , drop=FALSE] 
cor1_tissue = as.matrix(cor1_tissue)

row_names_cor1_t    = as.data.frame(strsplit(x=rownames(cor1_tissue),split="_vorinostat"))
colnames(cor1_tissue)[colnames(cor1_tissue)=="X786.0"] <- "786.0"
rownames (cor1_tissue) = as.data.frame (t(row_names_cor1_t[1,]))[,1]

names(cor1_tissue)[names(cor1_tissue)=="V1"] <- "Tissue"

cor1_tissue = as.matrix(cor1_tissue)
```



Tissue Annotation: Creating the function for the annotation

```{r}
cor1_tissue_an <- data.frame(V1 = rep(c("Renal", 
                                        "Lung", 
                                        "Breast", 
                                        "Leukemia", 
                                        "Colon", 
                                        "Prostate", 
                                        "Ovarian", 
                                        "Melanoma", 
                                        "CNS"
                                        ),
                                      c(8,9,6,5,7,2,7,9,6)
                                      )
                             )

row.names(cor1_tissue_an) <- colnames(cor1_tab)
```

#Heatmap

```{r}
cor1_colour = list(V1 = c("Renal" = "#A0E862", 
                       "Lung" = "#29C0DE", 
                       "Breast" = "#005BF8", 
                       "Leukemia" = "#845EC2", 
                       "Colon" = "#CA009E", 
                       "Prostate" = "#FF88CB", 
                       "Ovarian" = "#F1903D", 
                       "Melanoma" = "#FFC75F", 
                       "CNS" = "#F9F871"
                       )
                   )


cor1 = pheatmap(cor1_tab,
                annotation_colors = cor1_colour,
                annotation_col = cor1_tissue_an,
                fontsize = 6.5,
                fontsize_row= 5, 
                fontsize_col = 6,
                gaps_col=50,
                info = TRUE
                )
```



#Heatmap with breaks

```{r}
pheatmap(cor1_tab,
         annotation_colors = cor1_colour,
         annotation_col = cor1_tissue_an,
         fontsize = 6.5,
         fontsize_row= 5, 
         fontsize_col = 6,
         gaps_col=50,
         info = TRUE,
         cutree_rows = 2,
         cutree_cols = 4)
```



#   2.2.2 Heatmap with 100 Biomarkers and 59 cell lines                                                       #####
This is how the heatmap looks making any modifications.
```{r}
pheatmap(cor1.2_tab)
```

# Scaling the rows

```{r}
cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}

cor1.2_tab_norm <- t(apply(cor1.2_tab, 1, cal_z_score))
pheatmap(cor1.2_tab_norm)
```

# Ploting a dendogram and cutting the tree

```{r}
cor1.2_hclust <- hclust(dist(cor1.2_tab), method = "complete")

as.dendrogram(cor1.2_hclust) %>%
  plot(horiz = TRUE)

cor1.2_hclust_tree <- cutree(tree = as.dendrogram(cor1.2_hclust), k = 2)
```


To which cluster do some of our genes belong?
```{r}
head(cor1.2_hclust_tree)
```

We can see that different genes belong to different clusters. The clusters to which the genes belong are different when the relationship between more biomarkers is explored.


#Tissue Annotation:

#Here we can use the same annotation functions as in Heatmap with 30 Biomarkers and 59 cell lines: cor1_tissue_an and cor1_colour.


#Heatmap for a 100 Biomarkers

```{r}
cor1.2 = pheatmap(cor1.2_tab,
                  annotation_colors = cor1_colour,
                  annotation_col = cor1_tissue_an,
                  fontsize = 6.5,
                  fontsize_row= 5, 
                  fontsize_col = 6,
                  gaps_col=50,
                  info = TRUE
                  )
```

#Heatmap with breaks

```{r}
pheatmap(cor1.2_tab,
         annotation_colors = cor1_colour,
         annotation_col = cor1_tissue_an,
         fontsize = 6.5,
         fontsize_row= 5, 
         fontsize_col = 6,
         gaps_col=50,
         info = TRUE,
         cutree_rows = 2,
         cutree_cols = 4)
```


##  2.3   Dendogram                                                                                           
#   2.3.1 Dendogram with 30 Biomarkers and 59 cell lines                                                      

## CELL LINES

# ggplot integration

```{r}
d_gg <- cor1_tab %>% dist %>% hclust %>% as.dendrogram %>%
  set("branches_k_color", k=2) %>% set("branches_lwd", c(1.5,1,1)) %>%
  set("branches_lty", c(1,1,3,1,1,2)) %>%
  set("labels_colors") %>% set("labels_cex", c(.5,0.8)) %>% 
  set("nodes_pch", 19) %>% set("nodes_col", c("orange", "black", "plum", NA),
                               main = "How Vorinostat biomarkers define the relatiosnhip between cell lines",
  )

plot(d_gg)
```


# Polar dendogram

```{r}
hc_pd <- hclust(dist(cor1_tab))
d_pd <- as.dendrogram(hc_pd)

d_pd <- d_pd %>% 
  color_branches(k=2) %>% 
  color_labels

par(mar = rep(0,4))
circlize_dendrogram(d_pd, labels_track_height = NA, dend_track_height = .3) 
```

# Retrieving hierachical clustering from the heatmap

```{r}
cor1_ret_hclust <- pheatmap(cor1_tab, silent = TRUE)

hc_hmd <- hclust(dist(cor1_tab), "ave")
cor1_ret_hclust$tree_row %>% 
  as.dendrogram() %>%
  plot(horiz = TRUE)

cor1_ret_hclust$tree_row = cor1_ret_hclust$tree_row %>% color_branches(k=2) %>% color_labels

par(mar = c(3,1,1,7))
plot(cor1_ret_hclust$tree_row, horiz  = TRUE)
```


## BIOMARKERS

# ggplot integration

```{r}
d_gg <- t(cor1_tab) %>% dist %>% hclust %>% as.dendrogram %>%
  set("branches_k_color", k=2) %>% set("branches_lwd", c(1.5,1,1)) %>%
  set("branches_lty", c(1,1,3,1,1,2)) %>%
  set("labels_colors") %>% set("labels_cex", c(.5,0.8)) %>% 
  set("nodes_pch", 19) %>% set("nodes_col", c("orange", "black", "plum", NA),
                               main = "How Vorinostat biomarkers define the relatiosnhip between cell lines",
  )

plot(d_gg)
```


# Polar dendogram

```{r}
hc_pd <- hclust(dist(t(cor1_tab)))
d_pd <- as.dendrogram(hc_pd)

d_pd <- d_pd %>% 
  color_branches(k=2) %>% 
  color_labels

par(mar = rep(0,4))
circlize_dendrogram(d_pd, labels_track_height = NA, dend_track_height = .3) 

```


# Retrieving hierachical clustering from the heatmap

```{r}
cor1_ret_hclust <- pheatmap(t(cor1_tab), silent = TRUE)

hc_hmd <- hclust(dist(t(cor1_tab)), "ave")
cor1_ret_hclust$tree_row %>% 
  as.dendrogram() %>%
  plot(horiz = TRUE)

cor1_ret_hclust$tree_row = cor1_ret_hclust$tree_row %>% color_branches(k=2) %>% color_labels

par(mar = c(3,1,1,7))
plot(cor1_ret_hclust$tree_row, horiz  = TRUE)
```



#   2.3.2 Dendogram with 100 Biomarkers and 59 cell lines                                                     

## CELL LINES

# ggplot integration

```{r}
d_gg <- cor1.2_tab %>% dist %>% hclust %>% as.dendrogram %>%
  set("branches_k_color", k=2) %>% set("branches_lwd", c(1.5,1,1)) %>%
  set("branches_lty", c(1,1,3,1,1,2)) %>%
  set("labels_colors") %>% set("labels_cex", c(.5,0.8)) %>% 
  set("nodes_pch", 19) %>% set("nodes_col", c("orange", "black", "plum", NA),
                               main = "How Vorinostat biomarkers define the relationship between cell lines",
  )

plot(d_gg)
```



# Retrieving hierachical clustering from the heatmap

```{r}
cor1.2_ret_hclust <- pheatmap(cor1.2_tab, silent = TRUE)

hc_hmd <- hclust(dist(cor1.2_tab), "ave")
cor1.2_ret_hclust$tree_row %>% 
  as.dendrogram() %>%
  plot(horiz = TRUE)

cor1_ret_hclust$tree_row = cor1.2_ret_hclust$tree_row %>% color_branches(k=2) %>% color_labels

par(mar = c(3,1,1,7))
plot(cor1.2_ret_hclust$tree_row, horiz  = TRUE)
```


## BIOMARKERS

# ggplot integration

```{r}
d_gg <- t(cor1.2_tab) %>% dist %>% hclust %>% as.dendrogram %>%
  set("branches_k_color", k=2) %>% set("branches_lwd", c(1.5,1,1)) %>%
  set("branches_lty", c(1,1,3,1,1,2)) %>%
  set("labels_colors") %>% set("labels_cex", c(.5,0.8)) %>% 
  set("nodes_pch", 19) %>% set("nodes_col", c("orange", "black", "plum", NA),
                               main = "How Vorinostat biomarkers define the relatiosnhip between cell lines",
  )

plot(d_gg)
```


# Polar dendogram

```{r}
hc_pd <- hclust(dist(t(cor1_tab)))
d_pd <- as.dendrogram(hc_pd)

d_pd <- d_pd %>% 
  color_branches(k=2) %>% 
  color_labels

par(mar = rep(0,4))
circlize_dendrogram(d_pd, labels_track_height = NA, dend_track_height = .3) 

```


# Retrieving hierachical clustering from the heatmap

```{r}
cor1_ret_hclust <- pheatmap(t(cor1_tab), silent = TRUE)

hc_hmd <- hclust(dist(t(cor1_tab)), "ave")
cor1_ret_hclust$tree_row %>% 
  as.dendrogram() %>%
  plot(horiz = TRUE)

cor1_ret_hclust$tree_row = cor1_ret_hclust$tree_row %>% color_branches(k=2) %>% color_labels

par(mar = c(3,1,1,7))
plot(cor1_ret_hclust$tree_row, horiz  = TRUE)
```




#   2.3.3 Comparison of dendograms with 100 Biomarkers and 30 Biomarkers                     

##  2.4   Correlogram and Scatter Plot for 30 Biomarkers and 59 cell lines                                    

# Correlogram
```{r}
corrgram(cor1_tab, order=NULL, panel=panel.shade, text.panel=panel.txt,
         main="Correlogram of Biomarkers and Cell lines") 
```


# Scatter plots

Biomarkers 1 to 10
```{r}
pairs(cor1_tab, horInd = 1:10, verInd = 1:10, col="#1E90FF")
```

Biomarkers 10 to 20
```{r}
pairs(cor1_tab, horInd = 10:20, verInd = 10:20, col="#1E90FF")
```

Biomarkers 20 to 30
```{r}
pairs(cor1_tab, horInd = 20:30, verInd = 20:30, col="#1E90FF")
```


# Matrix for correlations: with 30 biomarkers and with 100 biomarkers

30 Biomarkers:
```{r}
cor = cor(cor1_tab)
heatmap(cor, col = cm.colors(256))
```

100 Biomarkers
```{r}
cor1.2 = cor(cor1.2_tab)
heatmap(cor1.2, col = cm.colors(256))
```


