---
title: "Specific"
author: "Laura Plutowski"
date: "20.06.2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
wd = ("/Users/laura.plutowski/Desktop/Uni/4.Semester/projekt/project-02-group-05")
``` 

```{r read_data, message = FALSE, warning = FALSE,include = FALSE}
library(readr)

Untreated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_untreated.rds"))
Treated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_treated.rds"))
Metadata = read.table(paste0(wd,"/data/NCI_TPW_metadata.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)
Cellline_annotation = read.table(paste0(wd,"/data/cellline_annotation.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)
Drug_annotation = read.table(paste0(wd,"/data/drug_annotation.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)


Treated <- as.data.frame(Treated)
Untreated <- as.data.frame(Untreated)
```

```{r, include = FALSE}
################## Creat Vorinostat data ####################

#Find cell lines, which belong to vorinostat:
#Untreated matrix:
UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))


#Same with treated matrix:
TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treated))


#Create Vorinostat Untreated matrix:
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]

#fold change matrix 
FC <- TreatedVorinostat - UntreatedVorinostat

#normalization 
#For normalization, a package needs to be installed: install.packages("BBmisc")
library(BBmisc)
FCnorm <- normalize(FC, method= "scale")
```

Now that we have determined the biomarkers via the T-test, we want to investigate their FC values, since our biomarkers should have the largest change in gene expression.

```{r include = FALSE}
### create data for p.values ###

#Create a common matrix for treated and untreated vorinostat data:
VorinostatTotal <- cbind(UntreatedVorinostat, TreatedVorinostat)

# find the needed colums 
 col_untreated = grep ('_0nM',colnames(VorinostatTotal))
 col_treated = grep ('_5000nM',colnames(VorinostatTotal))

pValues <- apply(VorinostatTotal, 1, function(x) t.test(x[col_untreated],x[col_treated],paired = TRUE, alternative = "two.sided")$p.value)

# sort the p.values 
sortedpValues <- sort(pValues, decreasing = FALSE)
sortedpValues <- as.matrix(sortedpValues)

# take the first 100 p.values for biomarkers 
biomarkers <- sortedpValues[1:100,]
biomarkers <- as.matrix(biomarkers)

# creat vector with genes 
biomarkers.genes = row.names(biomarkers)
# check if it has the correct length
length(biomarkers.genes)
```

```{r}
### create data for FC ###

FC <- TreatedVorinostat - UntreatedVorinostat

# work with mean of the rows because we only want to compare the genes 
FC_meanrow= rowMeans(FC)

# look at the absolute value, because genes could be also highly down regulated 
FC_meanrow=abs(FC_meanrow)

# sort the FC 
sortedFC <- sort(FC_meanrow, decreasing = FALSE)
sortedFC <- as.matrix(sortedFC)

# take the first 100 values like done above for compairison 
FC100<- sortedFC[1:100,]
FC100 <- as.matrix(FC100)

# creat vector with genes 
FC100.genes = row.names(FC100)
# check if it has the correct length
length(FC100.genes)


### compare genes vectors ### 

# Do we have exactly the same biomarkers?
setequal(FC100.genes,biomarkers.genes)

# How many of the biomarkers are different? 
diff= setdiff(FC100.genes,biomarkers.genes)
length(diff)

### compare p.Values and FC values in a new matrix ###

#creat new matrix with both informations
pV_FC= cbind(FC_meanrow,sortedpValues)
head(pV_FC)

```

Now we know that p.Value and FC value do not deliver exactly the same biomarkers. In fact, they differ by 100%. 
If we compare the values directly, we see that a low p.value does not correlate with a high FC value. This is because small values are also assigned small p.values by the T-test and the result of the test is therefore falsified. For this reason, we cannot trust the biomarkers from the T-test and will now determine them via the fold change values.

```{r}
### creat FC data ###
FC <- TreatedVorinostat - UntreatedVorinostat

# work with mean of the rows because we only want to compare the genes 
FC_meanrow= rowMeans(FC)


### sort the data ###

# work with absolute value to find the highest values
# because we want to have the most up and down regulated genes 
FC_abs= abs(FC_meanrow)

## sort the values to get the 100 largest values 

sortedFC_abs <- sort(FC_abs, decreasing = TRUE)
sortedFC_abs <- as.matrix(sortedFC_abs)

# take the first 100 for biomarkers 
biomarkers_FC = sortedFC_abs[1:100,]
biomarkers_FC <- as.matrix(biomarkers_FC)
head(biomarkers_FC)

# creat vector with gene names 
biomarkers_FC_genes= row.names(biomarkers_FC)
```

Now we have determined trusting biomarkers via the fold change. But with these biomarkers it is not taken into account that up- or down-regulating effects can cancel each other out. So now we're going to determine the biomarkers by their absolute FC values.

```{r}

# creat FC matrix
FCVorinostat <- TreatedVorinostat - UntreatedVorinostat

# work with mean of the rows because we only want to compare the genes:
FCVorinostatabs= abs(FCVorinostat)
FCVorinostatmean <- apply(FCVorinostatabs, 1, mean)

# sort the values to get the 100 largest values:
sortedgeneralbiomarker <- sort(FCVorinostatmean, decreasing = TRUE)
sortedgeneralbiomarker <- as.matrix(sortedgeneralbiomarker)

#select the top 100 general biomarkers:
top100generalbiomarkers = sortedgeneralbiomarker[1:100,]
top100generalbiomarkers <- as.matrix(top100generalbiomarkers)
head(top100generalbiomarkers)

#create vector with gene names:
generalbiomarkergenes = row.names(top100generalbiomarkers)
```

Now we want to compare the biomarkers with and without using the absolute value for determining to see how much they differ and to decide which genes we will continue to work with.

```{r}
### compairison ###

# Do we have exactly the same biomarkers?
setequal(biomarkers_FC_genes,generalbiomarkergenes)

# How many of the biomarkers are different? 

# biomarkers in biomarkers_FC_genes but not in generalbiomarkergenes
diff1= setdiff(biomarkers_FC_genes,generalbiomarkergenes)
length(diff1)
diff1

# biomarkers in generalbiomarkergenes but not in biomarkers_FC_genes
diff2= setdiff(generalbiomarkergenes,biomarkers_FC_genes)
length(diff2)
diff2
```

We see that only 5 genes differ in the biomarkers. Nevertheless, it is better to trust the absolute values, as they indicate the largest change in gene expression.
In order to be able to work better with the biomarkers, we creat a matrix with the information whether the genes are up or down regulated.  

``` {r include = FALSE}
#install.packages("knitr")
library(knitr)
```

```{r results = "asis"}
### creat matrix
FCVorinostatmean_noabs= rowMeans(FCVorinostat)

#Matrix with information about general up- or downregulation:
FCVorinostatmean_noabs_matrix <- as.matrix(FCVorinostatmean_noabs)

#loop, which adds those biomarkergenes to a vector, which are mostly upregulated:
i=1
Upregulated <- c()
while(i<=100) {
  if(FCVorinostatmean_noabs_matrix[generalbiomarkergenes[i],1] >0) {
    Upregulated <- union(Upregulated, generalbiomarkergenes[i])
  }
  i=i+1}

#Create a vector that includes for each gene sequentially if it is up- or downregulated:
i=1
Generalchange <- c()
while(i<=100) {
  if(FCVorinostatmean_noabs_matrix[generalbiomarkergenes[i],1] >0) {
    Generalchange <- append(Generalchange, "Up")
  } else {
    Generalchange <- append(Generalchange, "Down")
  }
  i=i+1}

#Bind the information about up/downregulation as a new column to the biomarkermatrix:
top100generalbiomarkers_withUporDown <- cbind(top100generalbiomarkers, Generalchange)
colnames(top100generalbiomarkers_withUporDown)[1] <- "FCmean"

knitr::kable(head(top100generalbiomarkers_withUporDown), caption = "Gene Expression Biomarkers")
```

Our biomarkers from the real FC values are visualized below, as they better reflect the change in gene expression: 

```{r plot_data, fig.height= 5, fig.width= 7}
### creat matrix with FC values (positive and negativ) 

# add the abs values to FC matrix 
FC_both= cbind(FC_meanrow,FC_abs)
FC_both=as.data.frame(FC_both)

# order this matrix 
FC_both_sorted <- FC_both[order(FC_both$FC_abs, decreasing = TRUE),]

# FC values of biomarkers
# take the first 100 of the sorted matrix, should be the same as un biomarkers_FC 
biomarkers_FC_values = FC_both_sorted[1:100,]
# remove the absolute values
biomarkers_FC_values <- subset( biomarkers_FC_values, select = -FC_abs)
biomarkers_FC_values = as.data.frame(biomarkers_FC_values)

par(mar=c(5, 4, 5, 9))
### visualization in bar plot 


# creat boxplot
barplot( height= biomarkers_FC_values$FC_meanrow, 
         main= "        Gene Expression Change by biomarkers", ylab="mean Fold change", xlab="biomarkers")

```
We see that most values are very similar to each other and that we have about the same number of genes regulated up and down.

Now we want to further investigate our biomarkers. First, we will see whether the targets for vorinostat in the metadata can also be found in our biomarkers.

```{r}
# search for HDAC in biomarkers 
target <- grep(pattern = "HDAC",colnames(top100generalbiomarkers))
length(target)

# In drug annotation HDAC genes are mentioned as target of vorinostat, but these genes are not in our defined 
# biomarkers, so know we want to check their FC values 

###find target of vorinostat 

target=Drug_annotation$target
target=as.data.frame(target)

# take row with vorinostat 
target_vorinostat=target[9,]
target_vorinostat

# new vector with names as strings, because otherwise problem with spacer | 
target_vorinostat=c("HDAC1","HDAC10","HDAC11","HDAC2","HDAC3","HDAC5","HDAC6","HDAC8","HDAC9")
```

In drug annotation HDAC genes are mentioned as target of vorinostat, but these genes are not in our defined biomarkers, so know we want to check their FC values.  

```{r}
### find FC values of targets 

# creat FC data 
FC =TreatedVorinostat - UntreatedVorinostat
FC= rowMeans(FC)
FC=as.data.frame(FC)
genes=row.names(FC)
FC_new=cbind(FC,genes)

```

``` {r, include = FALSE}
# filter data 
library(dplyr)
```

```{r}
FC_target=as.data.frame(filter(FC_new, genes %in% target_vorinostat))

knitr::kable(head(FC_target), caption = "FC values vorinostat targets")
```

The changes in gene expression are really low. Most of the HDAC genes are up regulated insted of being down regulated. This could be because the cell wants to compensate the inhibition of the enzymes by their increased formation. 