---
title: "Linear and Multiple Regressions"
author: "Florencia Zúñiga"
date: "7/9/2019"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```






```{r read_data, message = FALSE, warning = FALSE, include=FALSE}

### Loading packages

library(readr)
library(rstudioapi)
library(lattice)
library(e1071)
library(ggplot2)
library(scatterplot3d)
library(car)
library(scatterD3)
library(rgl)
library(dplyr)

wd = ("/GitHub/project-02-group-05")


### Reading the data
Untreated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_untreated.rds"))
Treated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_treated.rds"))

Metadata = read.table(paste0(wd,"/data/NCI_TPW_metadata.tsv"), 
                      header = TRUE, sep ="\t", stringsAsFactors = TRUE)

Sensitivity <- readRDS(paste0(wd,"/data/NegLogGI50.rds"))


Basal <- readRDS(paste0(wd,"/data/CCLE_basalexpression.rds"))
Copynumber <- readRDS(paste0(wd,"/data/CCLE_copynumber.rds"))
Mutations <- readRDS(paste0(wd,"/data/CCLE_mutations.rds"))

Metadata = read.table(paste0(wd,"/data/NCI_TPW_metadata.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)

Cellline_annotation = read.table(paste0(wd,"/data/cellline_annotation.tsv"), 
                                 header = TRUE, sep ="\t", stringsAsFactors = TRUE)
Drug_annotation = read.table(paste0(wd,"/data/drug_annotation.tsv"), 
                             header = TRUE, sep ="\t", stringsAsFactors = TRUE)

### Transforming the data

Treated <- as.data.frame(Treated)
Untreated <- as.data.frame(Untreated)
Sensitivity<- as.data.frame(Sensitivity)

### Data normalization

Untreated_norm <- apply(Untreated, 2, function(x){
  (x - mean(x)) / sd(x)
 })


Treated_norm <- apply(Treated, 2, function(x){
  (x - mean(x)) / sd(x)
 })


FC <- Treated - Untreated
FC_norm <- apply(FC, 2, function(x){
  (x - mean(x)) / sd(x)
 })

### Loading data of the biomarkers                                                                                                    
##  (1)  Creating Vorinostat

#Untreated matrix
UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))

#Same with treated matrix
TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treated))

#Define Vorinostat-data: 
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]

#fold change matrix 
FC <- TreatedVorinostat - UntreatedVorinostat

#Sensitivity 
vorinostat_Sensitivity_alleZeilen= grep ('vorinostat', rownames(Sensitivity))
vorinostat_Sensitivity_data= Sensitivity[vorinostat_Sensitivity_alleZeilen,]

##  (2)  Creating FC Data -  Finding the Biomarkers

FC <- TreatedVorinostat - UntreatedVorinostat

#We work with mean of the rows because we only want to compare the genes 
FC_meanrow= rowMeans(FC)

## Sorting the data 
#We work with absolute value to find the highest values, 
#because we want to have the most up and down regulated genes.
FC_abs= abs(FC_meanrow)

#We sort the values to get the 100 largest values 
sortedFC_abs <- sort(FC_abs, decreasing = TRUE)
sortedFC_abs <- as.matrix(sortedFC_abs)

#We select the first n for biomarkers 
biomarkers_FC30 = sortedFC_abs[1:30,]
biomarkers_FC30 <- as.matrix(biomarkers_FC30)

biomarkers_FC100 = sortedFC_abs[1:100,]
biomarkers_FC100 <- as.matrix(biomarkers_FC100)

## Creating a matrix with FC values, that are both positive and negative 
FC_both= cbind(FC_meanrow,FC_abs)
FC_both=as.data.frame(FC_both)

#Ordering this matrix 
FC_both_sorted <- FC_both[order(FC_both$FC_abs, decreasing = TRUE),]

#FC values of biomarkers: We select the first 100 of the sorted matrix. 
biomarkers_FC_values30 = FC_both_sorted[1:30,]

biomarkers_FC_values100 = FC_both_sorted[1:100,]

#Removing the absolute values
biomarkers_FC_values30 <- subset( biomarkers_FC_values30, select = -FC_abs)
biomarkers_FC_values30 = as.matrix(biomarkers_FC_values30)

biomarkers_FC_values100 <- subset( biomarkers_FC_values100, select = -FC_abs)
biomarkers_FC_values100 = as.matrix(biomarkers_FC_values100)

```


```{r, include=FALSE}
#### (4) Table 4: Drug sensitivity 
drug_sensitivity <- Sensitivity[-c(1:14),] 

drug_sensitivity = t(drug_sensitivity)

### (1)  Selection of 100 Biomarkers in copynumber 
BM_copynumber = Copynumber[ which(row.names(Copynumber) 
                                  %in% rownames(biomarkers_FC_values100)), ]



### Mean of each column
BM_Copynumber_meancol= colMeans(BM_copynumber)

```



```{r, include=FALSE}

CN = as.data.frame(BM_Copynumber_meancol)


```





```{r, include=FALSE}

#### (2) Table 2: All genes in copynumber 
CN_meancol= colMeans(Copynumber)

CN_all = as.data.frame(CN_meancol)

#### (3) Table 3: Selection of Doubling time from cellline_annotation

#Selecting the desired columns
Doubling_Time <- Cellline_annotation %>% 
  select(Cell_Line_Name, Doubling_Time)

#Changig the name of the "name" column to the names of the cell lines

row.names(Doubling_Time) <- Doubling_Time$Cell_Line_Name
Doubling_Time[1] <- NULL

```

***
##  1.  SIMPLE LINEAR REGRESSION WITH 100 BIOMARKERS: Drug sensitivity with copynumber of 100 biomarkers* 
***


```{r, include=FALSE}
### Data frames

CN = as.data.frame(BM_Copynumber_meancol)

DS = as.data.frame(drug_sensitivity)
```



```{r, include=FALSE}
#### Table with drug sensitivity and copynumber values per cell line



lm_tab2 = transform(merge(CN,DS,by=0,all=TRUE), row.names=Row.names, Row.names=NULL)


names(lm_tab2)[names(lm_tab2) == "BM_Copynumber_meancol"] <- "Copynumber_100_biomarkers"
names(lm_tab2)[names(lm_tab2) == "vorinostat"] <- "Drug_Sensitivity"


lm_tab2 <- na.omit(lm_tab2)
```


###  1.1 Predicting the fittness of the model      
<p>
<p>**Visualizations of copynumber values vs drug sensitivity**
<p>
#### (1) Scatter Plot

```{r, fig.align= 'center', echo=FALSE}
scatter.smooth(lm_tab2$Drug_Sensitivity, 
              lm_tab2$Copynumber_100_biomarkers, 
              col = "dodgerblue1",
              main = "Drug sensitivity & Copynumber Regression",
              xlab = "Drug sensitivity",
              ylab = "Copynumber",
              cex = 1.3,
              pch = 1)
```


#### (2) Checking for correlation

```{r}
cor(lm_tab2$Drug_Sensitivity, lm_tab2$Copynumber_100_biomarkers)
```

***
###   1.2 Linear Regression 

#### Linear Regression (y = slope * x + intercept)

```{r}
reg2 <- lm(Drug_Sensitivity ~ Copynumber_100_biomarkers, data = lm_tab2)
```


#### Details about the linear regression

```{r}
summary(reg2)
```
***
#### **y = slope * x + intercept** 
<p>
<p>

<p>

#### Drug_Sensitivity = 1.11 * x + 5.95
<p>
***
###   1.3 Visualization of regression   
<p>
<p>**Visualization of regression with confidence interval**
<p>
```{r, fig.align= 'center', echo=FALSE}
ggplot(data = lm_tab2, aes(x = Copynumber_100_biomarkers, y = Drug_Sensitivity)) +
  geom_point() +
  stat_smooth(method = "lm", col = "slateblue3", level = 0.975) +
  theme(panel.background = element_rect(fill = "white"),
        axis.line.x=element_line(),
        axis.line.y=element_line()) +
  ggtitle("Linear Model Fitted to Data with confidence interval in grey") 
```


                             
<p>
<p>**Residual diagnostics**
<p>

```{r, fig.align= 'center', echo=FALSE}
par(mar = c(4, 4, 2, 2), mfrow = c(1, 2))
plot(reg2, which = c(1, 2), col = "dodgerblue1")
```

***
##  2.  MULTIPLE REGRESSION WITH 100 BIOMARKERS: Drug sensitivity with doubling time and copynumber of 100 biomarkers 
***


```{r, include=FALSE}
### Data frames
CN = as.data.frame(BM_Copynumber_meancol)

DT = as.data.frame(Doubling_Time)

DS = as.data.frame(drug_sensitivity)

### Table with drug sensitivity and doubling time per cell line

lm_tab = transform(merge(DT,DS,by=0,all=TRUE), row.names=Row.names, Row.names=NULL)



### Table with drug sensitivity, copynumber and doubling time per cell line

lm_tab_m = transform(merge(CN, 
                           lm_tab,by=0,all=TRUE), 
                          row.names=Row.names, 
                          Row.names=NULL
                    )


names(lm_tab_m)[names(lm_tab_m) == "BM_Copynumber_meancol"] <- "Copynumber"
names(lm_tab_m)[names(lm_tab_m) == "vorinostat"] <- "Drug_Sensitivity"


lm_tab_m <- na.omit(lm_tab_m)
```


###   2.1 Predicting the fittness of the model       
<p>
<p>**Visualization of drug sensitivity vs copynumber and doubling time** 
<p>

#### (1) Sactter plot

```{r, fig.align= 'center', echo=FALSE}
plot(lm_tab_m , 
     pch=20 , 
     cex=1.5 , 
     col=rgb(0.5, 0.8, 0.9, 0.7)
    )
```

#### (2) Checking for correlation

```{r, echo=FALSE, fig.align= 'center'}
cor(lm_tab_m)
```


***
###   2.2 Multiple Regression                                                                                                

#### Multiple Regression

```{r}
reg_m <- lm(Drug_Sensitivity ~ Copynumber + Doubling_Time, data = lm_tab_m)
```


#### Details about the linear regression

```{r}
summary(reg_m)
```

***
###   2.3 Visualization of regression 
<p>
<p>**3D Plot with Regression Plane**
<p>
```{r, fig.align= 'center', echo=FALSE}
#Creating 3D plot
m_3d <- scatterplot3d(lm_tab_m, 
                     type = "h", 
                     color = "hotpink",
                     angle=55, 
                     pch = 16)
#Adding regression plane
reg_m_3D <- lm(Drug_Sensitivity ~ Copynumber + Doubling_Time, data = lm_tab_m)
m_3d$plane3d(reg_m_3D)

```
<p>
<p>**Residual diagnostics**
<p>
```{r, echo=FALSE, fig.align= 'center'}
par(mar = c(4, 4, 2, 2), mfrow = c(1, 2))
plot(reg_m, which = c(1, 2), col = "orchid1")
```

