---
title: "Functional Analysis of Biomarkers"
author: "Laura Plutowski and Franziska Heinkele"
date: "25 July 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
#wd = ("C:/Users/franz/Documents/GitHub/project-02-group-05")
wd = ("/Users/laura.plutowski/Desktop/Uni/4.Semester/projekt/project-02-group-05")
``` 


```{r read_data, message = FALSE, warning = FALSE,include = FALSE}
library(readr)
Untreated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_untreated.rds"))
Treated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_treated.rds"))
```



```{r, include = FALSE}
#Find cell lines, which belong to vorinostat:
UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))
TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treated))
#Create Vorinostat matrix:
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]
# Create Vorinostat Fold change matrix 
FC <- TreatedVorinostat - UntreatedVorinostat
# work with mean of the rows because we only want to compare the genes:
FCVorinostatabs= abs(FC)
FCVorinostatmean <- apply(FCVorinostatabs, 1, mean)
# sort the values to get the 100 largest values:
sortedgeneralbiomarker <- sort(FCVorinostatmean, decreasing = TRUE)
sortedgeneralbiomarker <- as.matrix(sortedgeneralbiomarker)
#select the top 100 general biomarkers:
top100generalbiomarkers = sortedgeneralbiomarker[1:100,]
top100generalbiomarkers <- as.matrix(top100generalbiomarkers)
#create vector with gene names:
generalbiomarkergenes = row.names(top100generalbiomarkers)
```


***
***

## 1. Functional Analysis "by hand" {#anchor1}


### 1.1. Create a table with biomarker features {#anchor2}


```{r}
vorinostat_annotation = read.csv2(paste0(wd,"/data/Biomarkers.csv"),header = TRUE, quote="\"")

```

```{r, include=FALSE}
tissue = vorinostat_annotation$affected.Tissue..if.specific.
general.function = vorinostat_annotation$general.Function
# rename 
names(vorinostat_annotation)[names(vorinostat_annotation)=="X...Gene"] <- "Gene" 
names(vorinostat_annotation)[names(vorinostat_annotation)=="affected.Tissue..if.specific."] <- "affected Tissue (if specific)" 

```

```{r, echo=FALSE}
knitr::kable(head(vorinostat_annotation, caption = "vorinostat annotation"))
```

***

### 1.2. Barplots of biomarker tissue and function {#anchor3}


```{r, fig.height= 5, fig.width= 7,echo = FALSE}
col=palette(rainbow(9))
par(mar=c(5, 4, 5, 9))
barplot(table(tissue), ylab="counts", main="Affected tissues by biomarkers", col=col, las=3, cex.names = 0.7)
```


***
***

## 2. Functional Analysis using "cluster profiler" {#anchor4}


### 2.1. Gene Ontology Analysis {#anchor5}


```{r,include=FALSE }
library(clusterProfiler)
library(org.Hs.eg.db)
```


```{r,warning=FALSE,message=FALSE,include=FALSE }
translated.genes= bitr(generalbiomarkergenes,
                        fromType="SYMBOL", 
                        toType="ENTREZID",
                        OrgDb = org.Hs.eg.db) 
head(translated.genes)
```


```{r,include=FALSE}
# load the needed library
library(DOSE)
```

```{r,include=FALSE}
# take the needed gene ID
gene=translated.genes$ENTREZID
head(gene)
```


***



```{r,warning=FALSE,message=FALSE,include=FALSE}
gene.df <- bitr(generalbiomarkergenes, fromType = "SYMBOL",
                toType = c("ENSEMBL"),
                OrgDb = org.Hs.eg.db)
```





__Enrichment Analysis Gene Ontology__ 
```{r,warning=FALSE,message=FALSE}
##  Cellular Component
ego.cc <- enrichGO(gene        = gene.df$ENSEMBL,
                 OrgDb         = org.Hs.eg.db,
                 keyType       = 'ENSEMBL',
                 ont           = "CC",
                 pAdjustMethod = "BH",
                 pvalueCutoff  = 0.01,
                 qvalueCutoff  = 0.05)
```

***

__Enrichment Analysis Gene Ontology of Biomakers according Cellular Component__
```{r,fig.height= 5, fig.width= 7, echo=FALSE}
# Dot Plot 
dotplot(ego.cc, showCategory=12)
```


***
***

### 2.2. KEGG Pathway Analysis {#anchor7}


```{r,warning=FALSE, results='hide',message=FALSE, include=FALSE}
library(KEGG.db)
library(org.Hs.eg.db)
library(enrichplot)
library(DOSE)
```


````{r,warning=FALSE,message=FALSE, include=FALSE}
gene.df <- bitr(generalbiomarkergenes, fromType = "SYMBOL",
                toType = c("ENTREZID"),
                OrgDb = org.Hs.eg.db)
```

```{r,warning=FALSE,message=FALSE, include=FALSE}
kk <- enrichKEGG(gene=gene.df$ENTREZID,
                 pvalueCutoff = 0.05,
                 pAdjustMethod = "none")
head(summary(kk))
```
***
__Biomarker Categories by KEGG__


```{r, fig.height= 5, fig.width= 7, echo=FALSE}
par(mfrow=c(2,1))
barplot(kk,showCategory=12)
```

***
***

### 2.3. Reactome Pathway Analysis {#anchor6}


```{r,warning=FALSE, results='hide',message=FALSE, include=FALSE}
library(ReactomePA)
```

```{r, warning=FALSE,message=FALSE,include=FALSE}
gene.df <- bitr(generalbiomarkergenes, fromType = "SYMBOL",
                toType = c("ENTREZID"),
                OrgDb = org.Hs.eg.db)
```

```{r,warning=FALSE,message=FALSE, include=FALSE}
x <- enrichPathway(gene=gene.df$ENTREZID,
                   pvalueCutoff = 0.05,
                   readable = T )
```

***

__Results from Reactome Pathway Analysis with Biomarkers__
```{r, fig.height= 5, fig.width= 7, echo=FALSE}
barplot(x,showCategory=12)
```

***

__Connection between Reactome Pathways of Biomarkers (colored by Fold Change)__
```{r, fig.height= 5, fig.width= 7, echo=FALSE}
# FC data for coloring 
FC=TreatedVorinostat-UntreatedVorinostat
FC<-FC[generalbiomarkergenes,]
names(FC) <- gene.df$ENTREZID
cnetplot(x,categorySize="geneNum",foldChange = FC)
```

***
***





