---
title: "Specific Analysis"
author: "Laura Plutowski, Franziska Heinkele and Mert Mayadag"
date: "25 July 2019"
output:
  html_document: default
  pdf_document: default
---

```{r, include = FALSE}
# Franzi
#wd = ("C:/Users/franz/Documents/GitHub/project-02-group-05/Broad Analysis")

# Laura 
wd = ("/Users/laura.plutowski/Desktop/Uni/4.Semester/projekt/project-02-group-05")
``` 


```{r read_data, message = FALSE, warning = FALSE, include=FALSE}
library(readr)
  Untreated = readRDS(paste0(wd,"/data/NCI_TPW_gep_untreated.rds"))
  Treated = readRDS(paste0(wd,"/data/NCI_TPW_gep_treated.rds"))
  Metadata = read_tsv(paste0(wd,"/data/NCI_TPW_metadata.tsv"))
  Drug_annotation = read.table(paste0(wd,"/data/drug_annotation.tsv"),
                               header = TRUE, sep ="\t", stringsAsFactors = TRUE)
    
```

***
# SPECIFIC ANALYSIS 
***

__Mechanism of vorinostat__ 
```{r, echo=FALSE}
Drug_annotation$Mechanism[grep(Drug_annotation$Drug, pattern= "vorinostat")]
```

***
__Target of vorinostat__ 
```{r,echo=FALSE}
target=Drug_annotation$target[grep(Drug_annotation$Drug, pattern= "vorinostat")]
target
```

***
__Cancer type vorinostat__ 
```{r,echo=FALSE}
Drug_annotation$indication[grep(Drug_annotation$Drug, pattern= "vorinostat")]
```



#Table of content

[1. General visualizations](#anchor1)

* [1.1. Pie chart of tissue types](#anchor2)

* [1.2.Density plot of general gene expression](#anchor3)

[2. Creation of Vorinostat-FC-matrix](#anchor4)

[3. Finding biomarkers](#anchor5)

* [3.1. Biomarkers received from two-sided T-test](#anchor6)

* [3.2. Biomarkers received from FC values](#anchor7)

       + [3.2.1. General Biomarkers](#anchor8)
       
       + [3.2.2. Up and Downregulation of biomarkers](#anchor9)

* [3.3. Biomarkers received from variance](#anchor10)
       
* [3.4. Comparison of the differently received biomarkers](#anchor11)

[4. Visualizations of biomarkers](#anchor12)

* [4.1. Density plot of biomarker gene expression](#anchor13)

* [4.2. Visualization of biomarkers with Vulcanoplot](#anchor15)

[5.Are the targets for vorinostat (HDACs) part of our biomarkers?](#anchor16)




***
## 1. General visualizations {#anchor1}

### 1.1. Pie chart of tissue types {#anchor2}

```{r, echo=FALSE }
Metadata <- as.matrix(Metadata)
Metadatatissue <- Metadata[1:819,"tissue"]
Treatedwithtissue <- rbind(Treated, Metadatatissue)

TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treatedwithtissue))
TreatedVorinostattissue <- Treatedwithtissue[,TreatedVorinostatcolumns]

#Create a pie chart:
    
palette(rainbow(9))
pie(table(TreatedVorinostattissue["Metadatatissue",]),
    main="Tissue types of Vorinostat-treated samples", 
    radius=1, 
    cex=0.8)
```


### 1.2.Density plot of general gene expression {#anchor3}


```{r,fig.height= 5, fig.width= 7, echo=FALSE}
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]

    UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]

par(mar=c(5, 4, 5, 9))
plot(density(UntreatedVorinostat), col="blue" ,xlab = "Gene expression",
     main = "Effects of Vorinostat on overall gene expression")
lines(density(TreatedVorinostat), col = "red")
legend("topright", inset = c(-0.4,0), legend=c("Untreated", "Treated") , xpd = TRUE, 
       pch=19, col = c("blue", "red"))
```


***

## 2. Create a Vorinostat-FC-matrix {#anchor4}

```{r,include=FALSE}
UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))
TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treated))
```


```{r,include=FALSE}
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]
```


```{r,include=FALSE}
# normalize the data 
UntreatedVorinostat <- apply(UntreatedVorinostat, 2, function(x){
  (x - mean(x)) / sd(x)
})


TreatedVorinostat <- apply(TreatedVorinostat, 2, function(x){
  (x - mean(x)) / sd(x)
})

FC <- TreatedVorinostat  - UntreatedVorinostat
```

***

## 3. Finding biomarkers {#anchor5}


## 3.1. Biomarkers received from two-sided paired T-test {#anchor6}

__Comparison of rows and columns of data frames Treated and Untreated__
```{r}
identical(rownames(UntreatedVorinostat), rownames(TreatedVorinostat))
```

```{r}
head(colnames(UntreatedVorinostat),3)
head(colnames(TreatedVorinostat),  3)

col_names_VU    = as.data.frame(strsplit(x=colnames(UntreatedVorinostat), split="_"))
col_names_VU    = as.data.frame (t(col_names_VU[1,]))[,1]

col_names_VT    = as.data.frame(strsplit(x=colnames(TreatedVorinostat),split="_"))
col_names_VT    = as.data.frame (t(col_names_VT[1,]))[,1]

identical(colnames(col_names_VU), colnames(col_names_VT))
```

__Calculation of significance using p-values__
```{r}
p_values_Vorinostat = sapply(1:13299, function(i){  
                      t.test( TreatedVorinostat[i,],UntreatedVorinostat[i,],paired = TRUE,
                                                        alternative = "two.sided")$p.value
})


p_values_Vorinostat           = as.data.frame (p_values_Vorinostat)
rownames(p_values_Vorinostat) = rownames((TreatedVorinostat))
colnames(p_values_Vorinostat) =  " p_values"

hist(p_values_Vorinostat[,1])
abline(v = summary(p_values_Vorinostat [,1]), col = c("blue", "red", "black", "orange"), 
    lty = 2)

Significance_Treated = sapply(1:13299, function(i) {
  # Return whether p-value  is < 0.05  
  p_values_Vorinostat [i,] < 0.05
})

p_values_Vorinostat$significance = Significance_Treated

p_values_Vorinostat = p_values_Vorinostat[ order(p_values_Vorinostat$` p_values`), ,drop = FALSE ]


p_values_Vorinostat[1:5,]
```
  
__Chosing 100 genes showing the smallest p-values__
```{r}
   biomarkers.pValues = head( p_values_Vorinostat,100 )
head(biomarkers.pValues)
```

```{r}
biomarkers.genes= rownames(biomarkers.pValues)
```

## 3.2. Biomarkers received from FC values {#anchor7}

### 3.2.1.General biomarkers {#anchor8}


__Find biomarkers using absolute FC Values__
```{r}
FC <- TreatedVorinostat - UntreatedVorinostat

FCVorinostatabs= abs(FC)
FCVorinostatmean <- apply(FCVorinostatabs, 1, mean)

#sort the values to get the 100 largest values:
sortedgeneralbiomarker <- sort(FCVorinostatmean, decreasing = TRUE)
sortedgeneralbiomarker <- as.matrix(sortedgeneralbiomarker)

#select the top 100 general biomarkers:
top100generalbiomarkers = sortedgeneralbiomarker[1:100,]
top100generalbiomarkers <- as.matrix(top100generalbiomarkers)

#Create a vector with gene names:
generalbiomarkergenes = row.names(top100generalbiomarkers)
```

```{r, echo=FALSE}
knitr::kable(head(top100generalbiomarkers), caption = "Top generalbiomarkers")
```





### 3.2.2. Up- and Downregulation of biomarkers {#anchor9}

__Biomarker without absolute values__
```{r}
FC_meanrow= rowMeans(FC)

# look at the absolute values:
FC_meanrow = abs(FC_meanrow)

# sort the FC 
sortedFC <- sort(FC_meanrow, decreasing = TRUE)
sortedFC <- as.matrix(sortedFC)

# take the first 100 values for comparison 
FC100<- sortedFC[1:100,]
FC100 <- as.matrix(FC100)

# create a vector with genenames 
biomarkergenes_noabs = row.names(FC100)
```

```{r, echo=FALSE}
knitr::kable(head(FC100), caption = "Top biomarkers without absolute values")
```


***

__Do we get the same biomarkers?__
```{r,echo=FALSE}
setequal(biomarkergenes_noabs,generalbiomarkergenes)
```


__How many of the biomarkers are different?__
```{r,echo=FALSE}
# biomarkers in biomarkergenes_noabs but not in generalbiomarkergenes
diff1= setdiff(biomarkergenes_noabs,generalbiomarkergenes)
length(diff1)
diff1
```


***
__Are the genes up- or down regulated?__
```{r}
FCVorinostatmean_noabs= rowMeans(FC)

#Matrix with information about general up- or downregulation:
FCVorinostatmean_noabs_matrix <- as.matrix(FCVorinostatmean_noabs)

#loop, which adds those biomarkergenes to a vector, which are mostly upregulated:
i=1
Upregulated <- c()
while(i<=100) {
  if(FCVorinostatmean_noabs_matrix[generalbiomarkergenes[i],1] >0) {
    Upregulated <- union(Upregulated, generalbiomarkergenes[i])
  }
  i=i+1}

#Create a vector that includes for each gene sequentially if it is up- or downregulated:
i=1
Generalchange <- c()
while(i<=100) {
  if(FCVorinostatmean_noabs_matrix[generalbiomarkergenes[i],1] >0) {
    Generalchange <- append(Generalchange, "Up")
  } else {
    Generalchange <- append(Generalchange, "Down")
  }
  i=i+1}

# Bind the information about up/downregulation as a new column to the biomarkermatrix:
top100generalbiomarkers_withUporDown <- cbind(top100generalbiomarkers, Generalchange)
# rename
colnames(top100generalbiomarkers_withUporDown)[1] <- "FCmean"
head(top100generalbiomarkers_withUporDown)
```

```{r,fig.height= 5, fig.width= 7, echo=FALSE}
top100generalbiomarkers_withUporDown=as.data.frame(top100generalbiomarkers_withUporDown)
change=top100generalbiomarkers_withUporDown$Generalchange
col=palette(rainbow(2))
barplot(table(change), ylab="counts", main="Number of up and down regulated genes", col=col)
```

### 3.3. Biomarkers received from variance {#anchor10}

__Selecting genes showing high variance__   
```{r, echo=TRUE}

Variances_Untreated             = as.data.frame (apply(UntreatedVorinostat,1,var))

colnames(Variances_Untreated)[1]= "var"
Most_Variances_Unt              = Variances_Untreated[ order(-Variances_Untreated$var), , drop= FALSE ] 

nintyfive_quartile_Untr         = quantile(Variances_Untreated$var, probs = 0.95)

Most_Variances_Unt              = Most_Variances_Unt[
                                 -which(Most_Variances_Unt$var < nintyfive_quartile_Untr),, drop= FALSE]
```

__Creating data frames from Treated and Untreated only with the selected genes__
```{r, echo=TRUE}
Highly_variant_Treated          = TreatedVorinostat  [ which(
                                  row.names(TreatedVorinostat) %in% rownames(Most_Variances_Unt)),]

Highly_variant_Untreated        = UntreatedVorinostat[ which(
                                  row.names(UntreatedVorinostat) %in% rownames(Most_Variances_Unt)),]
 ```
 
__Selecting genes with diminished variances after treatment__

```{r, echo=TRUE}

Variances_Untreated             = as.data.frame (apply(Highly_variant_Untreated,1,var  ))
Variances_Treated               = as.data.frame (apply(Highly_variant_Treated,1,var  ))

Variance_Change                 = as.data.frame (Variances_Treated[,1] - Variances_Untreated[,1])

row.names(Variance_Change)      = rownames(Variances_Treated)
colnames(Variance_Change)[1]    = "Var_Change"

lower_quartile_VC               = quantile (Variance_Change$Var_Change, probs = 0.05)
Most_neg_var_change             = Variance_Change[-which(
                                  Variance_Change$Var_Change > lower_quartile_VC),, drop= FALSE] 
                                  
Most_neg_var_change             = Most_neg_var_change[ 
                                  order(Most_neg_var_change$Var_Change), ,drop = FALSE ] 
 ```

 __Selecting 30 genes with highest negative variance change__
 ```{r, echo=TRUE}
Highest_VC                      = head(Most_neg_var_change, 30)


Highest_VC_Biomarkers           = FC[ which(row.names(FC) %in% rownames(Highest_VC)),]
colnames(Highest_VC_Biomarkers) = col_names_VT
 ```
__Do these genes show similar responses in a specific cancer type?__
 ```{r, echo=TRUE}
Metadata                        = as.data.frame(Metadata)
Metadata_Vorinostat             = Metadata    [ which(grepl  ("vorinostat" , Metadata$drug) ), ] 
Metadata_Vorinostat             = Metadata_Vorinostat [ order(Metadata_Vorinostat$cell), ,drop = FALSE ] 
Metadata_V_after                = subset(Metadata_Vorinostat, Metadata_Vorinostat$dose == "5000nM")
Metadata_V_after                = Metadata_V_after [, -which(colnames(Metadata_V_after) %in% 
                                                 c( "time", "drug", "sample", "dose" ))]
rownames(Metadata_V_after)      = Metadata_V_after$cell

colnames(Highest_VC_Biomarkers) = Metadata_V_after$tissue

heatmap(Highest_VC_Biomarkers)
```


### 3.4. Comparison of the differently received biomarkers {#anchor11}


__Do we have exactly the same biomarkers from pValues and FC?__
```{r,echo=FALSE}
setequal(generalbiomarkergenes,biomarkers.genes)
```

__How many of the biomarkers are different?__
```{r,echo=FALSE}
diff= setdiff(generalbiomarkergenes,biomarkers.genes)
length(diff)
```

***

``` {r include = FALSE}
library(knitr)
```

__Comparison of most significant p.Values and their FC values__
```{r,results='hide',message=FALSE, include=FALSE}
#sort FC matrix according biomarkers from pValues 
FC=TreatedVorinostat-UntreatedVorinostat
FC_m=rowMeans(FC)
FC_m=as.data.frame(FC_m)
FC =FC_m[biomarkers.genes,]
FC=as.data.frame(FC)
```

```{r, echo=FALSE}
# bind the matrices for Comparison
pV_FC= cbind(FC,biomarkers.pValues)
pV_FC=as.data.frame(pV_FC)
head(pV_FC)
```



***

#__TAKE THE BIOMARKERS FROM ABSOLUTE FC VALUES __

***
## 4. Visualizations of biomarkers {#anchor12}


### 4.1. Density plot of biomarker gene expression {#anchor13}


```{r,fig.height= 5, fig.width= 7, echo=FALSE}

Treated_Vorinostat_biomarker <- Treated[generalbiomarkergenes,TreatedVorinostatcolumns]
Untreated_Vorinostat_biomarker <- Untreated[generalbiomarkergenes,UntreatedVorinostatcolumns]

par(mar=c(5, 4, 5, 9))
plot(density(Untreated_Vorinostat_biomarker),col="blue" ,xlab = "Gene expression",
     main = "Effects of Vorinostat on biomarker gene expression")

lines(density(Treated_Vorinostat_biomarker), col = "red")

legend("topright", inset = c(-0.4,0), legend=c("Untreated", "Treated") , xpd = TRUE, 
       pch=19, col = c("blue", "red"))
```


***


### 4.2. Visualization of biomarkers with Vulcanoplot {#anchor15}


```{r, include=FALSE}
FC=TreatedVorinostat-UntreatedVorinostat
T.testFC <- apply(FC, 1, t.test)
```


```{r,include=FALSE}
TResults <- lapply(T.testFC, function(.tres) {      
  
  data.frame(t.value=.tres[1],dfs=.tres[2],conf.int1=.tres$conf.int[1],conf.int2=
               .tres$conf.int[2],p.value=.tres[3])
})

Tfinalresults <- do.call(rbind, TResults)
```


```{r message=FALSE, warning=FALSE,include=FALSE}
library(EnhancedVolcano)
```

```{r,include=FALSE}
log2FC <- apply(FC, 1, mean)
```

```{r,include=FALSE}
FCandpvalues <- cbind(Tfinalresults, log2FC)
```


```{r, fig.height= 5, fig.width= 7, echo=FALSE}
keyvals <- rep('black', nrow(FCandpvalues))
names(keyvals) <- rep('FC<Top100', nrow(FCandpvalues))

keyvals[which(FCandpvalues$log2FC >= min(FC100))] <- 'gold'
names(keyvals)[which(FCandpvalues$log2FC >= min(FC100))] <- 'positive FC in Top100'

keyvals[which(FCandpvalues$log2FC <= -min(FC100))] <- 'green'
names(keyvals)[which(FCandpvalues$log2FC <= -min(FC100))] <- 'negative FC in Top100'

EnhancedVolcano(FCandpvalues,
                lab = rownames(FCandpvalues),
                x = 'log2FC',
                y = 'p.value',
                title = 'Significance versus fold change of all genes',
                selectLab = biomarkergenes_noabs,
                transcriptLabSize = 1.8,
                colOverride = keyvals,
                pCutoff = 10e-14,
                FCcutoff = min(FC100)
              )
```






## 5. Are the targets for vorinostat (HDACs) part of our biomarkers? {#anchor16}

__What are the targets of vorinostat?__
```{r, echo=FALSE}
###find target of vorinostat 

target=Drug_annotation$target
target=as.data.frame(target)

# take row with vorinostat 
target_vorinostat=target[9,]
target_vorinostat

#create a vector with targetnames as strings

target_vorinostat=c("HDAC1","HDAC10","HDAC11","HDAC2","HDAC3","HDAC5","HDAC6",
                    "HDAC8","HDAC9")
```

***
__Do we have HDACs in our biomarkers?__
```{r, echo=FALSE}
# search for HDAC in biomarkers 
target <- grep(pattern = "HDAC",colnames(top100generalbiomarkers))
length(target)
```

***
__Which FC values do they have?__
```{r, include=FALSE}
### find FC values of targets 

# create FC data 
FC_meanrow = as.data.frame(FC_meanrow)
genes = row.names(FC_meanrow)
FC_new=cbind(FC_meanrow,genes)

# search for HDAC in biomarkers 
target <- grep(pattern = "HDAC",colnames(top100generalbiomarkers))
length(target)

library(dplyr)

# filter data 
FC_target = as.data.frame(filter(FC_new, genes %in% target_vorinostat))
FC_target
```

```{r, echo=FALSE}
knitr::kable(head(FC_target), caption = "FC values vorinostat targets")
```


***

```{r, include=FALSE}
first10biomarkers= generalbiomarkergenes[1:10]

c=union(target_vorinostat,first10biomarkers)
length(c)
```

```{r, include=FALSE}
FC= TreatedVorinostat-UntreatedVorinostat
FC=as.data.frame(FC)

FC_HDAC <-FC[c,]
```

```{r, include=FALSE}
FC_HDAC =na.omit(FC_HDAC)
```
***

```{r, fig.height= 5, fig.width= 7, echo=FALSE}
palette(rainbow(17))
boxplot(t(FC_HDAC), las=3, col = col, ylab="Fold Change Values", cex.axis= 0.65,  main="Comparison of Biomarker and target gene expression change ")
abline(v=7.5, col="blue", lty=5)
```

*** 

#__NEXT:  What are the functions of the biomarkers?__


