---
title: "Specfic Analysis"
author: "Laura Plutowski"
date: "20.06.2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
wd = ("/Users/laura.plutowski/Desktop/Uni/4.Semester/projekt/project-02-group-05")
``` 

```{r read_data, message = FALSE, warning = FALSE,include = FALSE}
library(readr)

Untreated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_untreated.rds"))
Treated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_treated.rds"))
Metadata = read.table(paste0(wd,"/data/NCI_TPW_metadata.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)
Cellline_annotation = read.table(paste0(wd,"/data/cellline_annotation.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)
Drug_annotation = read.table(paste0(wd,"/data/drug_annotation.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)


Treated <- as.data.frame(Treated)
Untreated <- as.data.frame(Untreated)
```

```{r, include = FALSE}
################## Creat Vorinostat data ####################

#Find cell lines, which belong to vorinostat:
#Untreated matrix:
UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))


#Same with treated matrix:
TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treated))


#Create Vorinostat Untreated matrix:
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]

#fold change matrix 
FC <- TreatedVorinostat - UntreatedVorinostat

#normalization 
#For normalization, a package needs to be installed: install.packages("BBmisc")
library(BBmisc)
FCnorm <- normalize(FC, method= "scale")
```

Now that we have determined the biomarkers via the T-test, we want to investigate their FC values, since our biomarkers should have the largest change in gene expression. 
```{r}
### create data for p.values ###

#Create a common matrix for treated and untreated vorinostat data:
VorinostatTotal <- cbind(UntreatedVorinostat, TreatedVorinostat)

# find the needed colums 
 col_untreated = grep ('_0nM',colnames(VorinostatTotal))
 col_treated = grep ('_5000nM',colnames(VorinostatTotal))

pValues <- apply(VorinostatTotal, 1, function(x) t.test(x[col_untreated],x[col_treated],paired = TRUE, alternative = "two.sided")$p.value)

# sort the p.values 
sortedpValues <- sort(pValues, decreasing = FALSE)
sortedpValues <- as.matrix(sortedpValues)

# take the first 100 p.values for biomarkers 
biomarkers <- sortedpValues[1:100,]
biomarkers <- as.matrix(biomarkers)

# creat vector with genes 
biomarkers.genes = row.names(biomarkers)
# check if it has the correct length
length(biomarkers.genes)



### create data for FC ###

FC <- TreatedVorinostat - UntreatedVorinostat

# work with mean of the rows because we only want to compare the genes 
FC_meanrow= rowMeans(FC)

# look at the absolute value, because genes could be also highly down regulated 
FC_meanrow=abs(FC_meanrow)

# sort the FC 
sortedFC <- sort(FC_meanrow, decreasing = FALSE)
sortedFC <- as.matrix(sortedFC)

# take the first 100 values like done above for compairison 
FC100<- sortedFC[1:100,]
FC100 <- as.matrix(FC100)

# creat vector with genes 
FC100.genes = row.names(FC100)
# check if it has the correct length
length(FC100.genes)


### compare genes vectors ### 

# Do we have exactly the same biomarkers?
setequal(FC100.genes,biomarkers.genes)

# How many of the biomarkers are different? 
diff= setdiff(FC100.genes,biomarkers.genes)
length(diff)

### compare p.Values and FC values in a new matrix ###

#creat new matrix with both informations
pV_FC= cbind(FC_meanrow,sortedpValues)
head(pV_FC)

```
Now we know that p.Value and FC value do not deliver exactly the same biomarkers. In fact, they differ by 100%. 
If we compare the values directly, we see that a low p.value does not correlate with a high FC value. This is because small values are also assigned small p.values by the T-test and the result of the test is therefore falsified. For this reason, we cannot trust the biomarkers from the T-test and will now determine them via the fold change values.To avoid that up- or down-regulating effects cancel each other out, we will work on the amount of biomarkers with the absolute value of FC values.

```{r}

# creat FC matrix
FCVorinostat <- TreatedVorinostat - UntreatedVorinostat

# work with mean of the rows because we only want to compare the genes:
FCVorinostatabs= abs(FCVorinostat)
FCVorinostatmean <- apply(FCVorinostatabs, 1, mean)

# sort the values to get the 100 largest values:
sortedgeneralbiomarker <- sort(FCVorinostatmean, decreasing = TRUE)
sortedgeneralbiomarker <- as.matrix(sortedgeneralbiomarker)

#select the top 100 general biomarkers:
top100generalbiomarkers = sortedgeneralbiomarker[1:100,]
top100generalbiomarkers <- as.matrix(top100generalbiomarkers)
head(top100generalbiomarkers)

#create vector with gene names:
generalbiomarkergenes = row.names(top100generalbiomarkers)
```

Now we have determined trusting biomarkers via the fold change. However, since we have worked with the amount, we cannot say whether our genes were up- or down-regulated. Therefore, we will now determine the biomarkers via the FC values for another time, but this time we will work without an absolute value. 


```{r}
### creat FC data ###
FC <- TreatedVorinostat - UntreatedVorinostat

# work with mean of the rows because we only want to compare the genes 
FC_meanrow= rowMeans(FC)


### sort the data ###

# work with absolute value to find the highest values
# because we want to have the most up and down regulated genes 
FC_abs= abs(FC_meanrow)

## sort the values to get the 100 largest values 

sortedFC_abs <- sort(FC_abs, decreasing = TRUE)
sortedFC_abs <- as.matrix(sortedFC_abs)

# take the first 100 for biomarkers 
biomarkers_FC = sortedFC_abs[1:100,]
biomarkers_FC <- as.matrix(biomarkers_FC)
head(biomarkers_FC)

# creat vector with gene names 
biomarkers_FC_genes= row.names(biomarkers_FC)
```

Now we want to compare the biomarkers with and without using the absolute value for determining to see how much they differ and to decide which genes we will continue to work with.

```{r}
### compairison ###

# Do we have exactly the same biomarkers?
setequal(biomarkers_FC_genes,generalbiomarkergenes)

# How many of the biomarkers are different? 

# biomarkers in biomarkers_FC_genes but not in generalbiomarkergenes
diff1= setdiff(biomarkers_FC_genes,generalbiomarkergenes)
length(diff1)
diff1

# biomarkers in generalbiomarkergenes but not in biomarkers_FC_genes
diff2= setdiff(generalbiomarkergenes,biomarkers_FC_genes)
length(diff2)
diff2
```

We see that only 5 genes differ in the biomarkers. Since the non-use of the absolute value gives us additional information as to whether our genes are generally regulated up or down, we will continue to work with these biomarkers (biomarkers_FC_genes).

In order to be able to work better with the biomarkers, we have created a matrix from the FC values of these genes. 

```{r}
### creat matrix with FC values (positive and negativ) 

# add the abs values to FC matrix 
FC_both= cbind(FC_meanrow,FC_abs)
FC_both=as.data.frame(FC_both)

# order this matrix 
FC_both_sorted <- FC_both[order(FC_both$FC_abs, decreasing = TRUE),]

# FC values of biomarkers
# take the first 100 of the sorted matrix, should be the same as un biomarkers_FC 
biomarkers_FC_values = FC_both_sorted[1:100,]
# remove the absolute values
biomarkers_FC_values <- subset( biomarkers_FC_values, select = -FC_abs)
# this is a matrix with mean of the FC values from all samples for each biomarker
biomarkers_FC_values = as.data.frame(biomarkers_FC_values)


### Creat Matirx with biomarkers with FC of all samples 
FCbiomarkers <- FC[biomarkers_FC_genes,]
```

Our biomarkers with their FC values are visualized below: 

```{r plot_data, fig.height= 5, fig.width= 7}
par(mar=c(5, 4, 5, 9))
### visualization in bar plot 
tissue <- Metadata$tissue
palette(rainbow(9))

# creat boxplot
barplot( height= biomarkers_FC_values$FC_meanrow,col = tissue, border = NA, 
         main= "        Gene Expression Change by biomarkers colored according tissues", ylab="mean Fold change", xlab="biomarkers")
levels <- as.factor(levels(tissue))
legend("topright", inset = c(-0.4,0.0), legend= levels(tissue), xpd = TRUE, pch=19, col = levels, title = "tissues")
```
