---
title: "Specific Analysis"
author: "Laura Plutowski and Franziska Heinkele"
date: "26 Juni 2019"
output:
  html_document: default
  pdf_document: default
---

```{r, include = FALSE}
# Franzi
#wd = ("C:/Users/franz/Documents/GitHub/project-02-group-05/Broad Analysis")

# Laura 
wd = ("/Users/laura.plutowski/Desktop/Uni/4.Semester/projekt/project-02-group-05")
``` 

Load data:
```{r read_data, message = FALSE, warning = FALSE}
library(readr)
  Untreated = readRDS(paste0(wd,"/data/NCI_TPW_gep_untreated.rds"))
  Treated = readRDS(paste0(wd,"/data/NCI_TPW_gep_treated.rds"))
  Metadata = read_tsv(paste0(wd,"/data/NCI_TPW_metadata.tsv"))
  Drug_annotation = read.table(paste0(wd,"/data/drug_annotation.tsv"),
                               header = TRUE, sep ="\t", stringsAsFactors = TRUE)
    
```

***
# SPECIFIC ANALYSIS 
***
<div style="text-align: justify">From now on, we decided to focus our analysis on Vorinostat.</div>

```{r}
Drug_annotation$Mechanism[grep(Drug_annotation$Drug, pattern= "vorinostat")]
```
The annotation data include information about vorinostats working mechanism as a __HDAC inhibitor__ (HDAC = histone deacetylases).
Normally, cells maintain a dynamic balance between acetylated and deacetylated histones.

* __Acetylated form__: Histones loose their positive charge when they are acetylated, what means that the negatively charged DNA is less bound to the histone proteins and the genes are more accessible. This form is promoted by the HDAC-Inhibitors, including vorinostat.

* __Deacetylated form__: Histones have a positive charge which causes a strong DNA binding. Genes can be hardly transcribed, including tumor-suppressorgenes. This form is promoted by the HDACs.

Vorinostats application as an anti-cancer medicine might origin from a better access to tumor-suppressorgenes in the acetylated form.

```{r}
target=Drug_annotation$target[grep(Drug_annotation$Drug, pattern= "vorinostat")]
```

```{r, echo=FALSE}
target
```
<div style="text-align: justify">
We see, that the targets of vorinostat are exclusively HDACs. In further analyses, we want to check their gene expression change under vorinostat- treatment since we can imagine that an increased inhibition might result in a higher production of HDACs to maintain the balance.
</div>
```{r}
Drug_annotation$indication[grep(Drug_annotation$Drug, pattern= "vorinostat")]
```
Until now, the T-cell lymphoma is the only cancer-type which indicates vorinostat-ingestion.


#Table of content

[1. General visualizations](#anchor1)

* [1.1. Pie chart of tissue types](#anchor2)

* [1.2.Density plot of general gene expression](#anchor3)

[2. Creation of Vorinostat-FC-matrix](#anchor4)

[3. Finding biomarkers](#anchor5)

* [3.1. Biomarkers received from two-sided T-test](#anchor6)

* [3.2. Biomarkers received from variance](#anchor10)

* [3.3. Biomarkers received from FC values](#anchor7)

       + [3.3.1. General Biomarkers](#anchor8)
       
       + [3.3.2. Up and Downregulation of biomarkers](#anchor9)
       
* [3.4. Comparison of the differently received biomarkers](#anchor11)

[4. Visualizations of biomarkers](#anchor12)

* [4.1. Density plot of biomarker gene expression](#anchor13)

* [4.2. Barplots of up- and downregulation of biomarkers](#anchor14)

* [4.3. Visualization of biomarkers with Vulcanoplot](#anchor15)

[5.Are the targets for vorinostat (HDACs) part of our biomarkers?](#anchor16)

[6.PCA with general-biomarker-matrix](#anchor17)

[7.Find biomarkers only for leukemia-celllines](#anchor18)



***
## 1. General visualizations {#anchor1}

### 1.1. Pie chart of tissue types {#anchor2}
<div style="text-align: justify">
Which tissue types are we dealing with when we work with the vorinostat-treated celllines? To get an overview of this, we use tissue information from Metadata and add it to the Treated-matrix, then we select those columns which belong to Vorinostat treatment:
</div>
```{r}
Metadata <- as.matrix(Metadata)
Metadatatissue <- Metadata[1:819,"tissue"]
Treatedwithtissue <- rbind(Treated, Metadatatissue)

TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treatedwithtissue))
TreatedVorinostattissue <- Treatedwithtissue[,TreatedVorinostatcolumns]

#Create a pie chart:
    
palette(rainbow(9))
pie(table(TreatedVorinostattissue["Metadatatissue",]),
    main="Tissue types of Vorinostat-treated samples", 
    radius=1, 
    cex=0.8)
```
<div style="text-align: justify">
We see, that most of our samples belong to the tissue types "lung", "melanoma" and "renal". However, this does not mean that Vorinostat works best there. This is only the selection of samples for the dataset.</div>

### 1.2.Density plot of general gene expression {#anchor3}
<div style="text-align: justify">
This plot is done to compare the gene expression of all provided genes with and without vorinostat treatment: </div>
```{r,fig.height= 5, fig.width= 7}
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]

    UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]

par(mar=c(5, 4, 5, 9))
plot(density(UntreatedVorinostat), col="blue" ,xlab = "Gene expression",
     main = "Effects of Vorinostat on overall gene expression")
lines(density(TreatedVorinostat), col = "red")
legend("topright", inset = c(-0.4,0), legend=c("Untreated", "Treated") , xpd = TRUE, 
       pch=19, col = c("blue", "red"))
```
<div style="text-align: justify">
Altough Vorinostats working mechanism as a HDAC inhibitor should lead to a better access to genes, we do not see a higher expression in terms of all genes. We are interested in how this plot changes when we plot only our biomarkers, so we are going to repeat the density plot later when we identified them. </div>

***

## 2. Create a Vorinostat-FC-matrix {#anchor4}
<div style="text-align: justify">
For the specific analysis, it is laborious to work with the whole dataset which includes data from all 15 drugs. Consequently, we reduce the data-sets by selecting those samples that belong to vorinostat: </div>

```{r}
UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))
TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treated))
```

We want to keep all genes but only the selected samples:
```{r}
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]
```

Moreover, we have to normalize the data to remove batch effects. 
```{r}
# normalize the data 
UntreatedVorinostat <- apply(UntreatedVorinostat, 2, function(x){
  (x - mean(x)) / sd(x)
})


TreatedVorinostat <- apply(TreatedVorinostat, 2, function(x){
  (x - mean(x)) / sd(x)
})

FC <- TreatedVorinostat  - UntreatedVorinostat
```

***

## 3. Finding biomarkers {#anchor5}
<div style="text-align: justify">
Biomarkers are the genes which respond to the treatment. 
In order to find these, we applied different methods and compared the results.
</div>

## 3.1. Biomarkers received from two-sided paired T-test (by Mert Mayadag){#anchor6}
<div style="text-align: justify">
In order to find the genes which show significant changes in expression levels after treatment, two-sided paired T-test was performed to calculate the p-values of gene expression changes.

Formulation of Hypotheses:
 __H0 hypothesis__: Gene expression does not change significantly by drug treatment.
 __H1 hypothesis__: Gene expression changes significantly by drug treatment.

For the application, the data frames "Treated" and "Untreated" should consist of same genes in rows and same cell lines in the sample names.

Are the genes, listed in rows, identical?
</div>
```{r}
identical(rownames(UntreatedVorinostat), rownames(TreatedVorinostat))
```
<div style="text-align: justify">
However, columns have different names as these include information about drug dosis. The comparison of sample names, which include cell line information, can show us, whether the same cell lines are used and if the order is identical.
</div>
```{r, echo= FALSE }
head(colnames(UntreatedVorinostat),2)
head(colnames(TreatedVorinostat),  2)

# We can split the names before the underline characters, save these informations as data frames and look if the cell line names of both sample groups are identical
col_names_VU    = as.data.frame(strsplit(x=colnames(UntreatedVorinostat), split="_"))
col_names_VU    = as.data.frame (t(col_names_VU[1,]))[,1]

col_names_VT    = as.data.frame(strsplit(x=colnames(TreatedVorinostat),split="_"))
col_names_VT    = as.data.frame (t(col_names_VT[1,]))[,1]

identical(colnames(col_names_VU), colnames(col_names_VT))
```

<div style="text-align: justify">
Now we can apply the t-test and observe the significance of changes by investigating p-values.
</div>
```{r}
p_values_Vorinostat = sapply(1:13299, function(i){  
                      t.test( TreatedVorinostat[i,],UntreatedVorinostat[i,],paired = TRUE,
                                                        alternative = "two.sided")$p.value
})
```


```{r, echo= FALSE}
p_values_Vorinostat           = as.data.frame (p_values_Vorinostat)
rownames(p_values_Vorinostat) = rownames((TreatedVorinostat))
colnames(p_values_Vorinostat) =  " p_values"

hist(p_values_Vorinostat[,1], main= "Distribution of P values", xlab= " P Values of Expression Changes")
abline(v = summary(p_values_Vorinostat [,1]), col = c("blue", "red", "black", "orange"), 
    lty = 2)

Significance_Treated = sapply(1:13299, function(i) {
  # Return whether p-value  is < 0.05  
  p_values_Vorinostat [i,] < 0.05
})

p_values_Vorinostat$significance = Significance_Treated

p_values_Vorinostat = p_values_Vorinostat[ order(p_values_Vorinostat$` p_values`), ,drop = FALSE ]


p_values_Vorinostat[1:5,]
```
<div style="text-align: justify">
A lot of values are essentially zero, which would mean that many genes have showed significant responses to the treatment. This is ,however, dubious. 
  
For now, we will select the 100 genes showing smallest p-values and compare these later with biomarkers found by other methods.
</div>
```{r, echo= FALSE}
biomarkers.pValues = head( p_values_Vorinostat,100 )

biomarkers.genes= rownames(biomarkers.pValues)
```

### 3.2. Biomarkers received from variance (by Mert Mayadag) {#anchor10}
<div style="text-align: justify">
In this part, we firstly selected the genes showing high variance among the 59 cell lines. 

We hypothesized that high variance of a gene could signify either 
- cell line specificity
or
- cancerogenic activity.

Then, we hypothesized that a decrease in variance would be more probable in cancerogenic genes, rather than cell line specific genes and therefore, selected the genes showing a negative variance decrease. 
</div>
The steps were: 

2.1 - Selecting genes with a variance above 95% quartile of variances before treatment
2.2 - Selecting genes which showed a significant decrease in their high variances
2.3 - Depicting the drug responses(FC) of these genes with a heatmap

```{r, echo=FALSE}

# 1- Selecting the upper 5% of genes with highest variances
Variances_Untreated             = as.data.frame (apply(UntreatedVorinostat,1,var))

colnames(Variances_Untreated)[1]= "var"
Most_Variances_Unt              = Variances_Untreated[ order(-Variances_Untreated$var), , drop= FALSE ] 

nintyfive_quartile_Untr         = quantile(Variances_Untreated$var, probs = 0.95)
```

```{r, echo=TRUE}
Most_Variances_Unt              = Most_Variances_Unt[
                                 -which(Most_Variances_Unt$var < nintyfive_quartile_Untr),, drop= FALSE]
```

```{r, echo=FALSE}
# 2 - Selecting genes which showed highest decreases in their high variance

# Creating smaller data frames from Treated and Untreated with only the selected genes
Highly_variant_Treated          = TreatedVorinostat  [ which(
                                  row.names(TreatedVorinostat) %in% rownames(Most_Variances_Unt)),]

Highly_variant_Untreated        = UntreatedVorinostat[ which(
                                  row.names(UntreatedVorinostat) %in% rownames(Most_Variances_Unt)),]
 
Variances_Untreated             = as.data.frame (apply(Highly_variant_Untreated,1,var  ))
Variances_Treated               = as.data.frame (apply(Highly_variant_Treated,1,var  ))

```

```{r, echo=TRUE}
Variance_Change                 = as.data.frame (Variances_Treated[,1] - Variances_Untreated[,1])
```

```{r, echo=FALSE}
row.names(Variance_Change)      = rownames(Variances_Treated)
colnames(Variance_Change)[1]    = "Var_Change"

lower_quartile_VC               = quantile (Variance_Change$Var_Change, probs = 0.05)
```

```{r, echo=TRUE}
Most_neg_var_change             = Variance_Change[-which(
                                  Variance_Change$Var_Change > lower_quartile_VC),, drop= FALSE] 
```

```{r, echo=FALSE}
# We focus on the negative part of the variance change, as we expect treatment to diminish the variance
Most_neg_var_change             = Most_neg_var_change[ 
                                  order(Most_neg_var_change$Var_Change), ,drop = FALSE ] 

# Here we select 30 genes which had a high variance at the beginning 
# and showed a variance decrease after treatment
Highest_VC                      = head(Most_neg_var_change, 30)

# We get for these genes the drug response (Fold Change) values for depicting with a heatmap
Highest_VC_Biomarkers           = FC[ which(row.names(FC) %in% rownames(Highest_VC)),]
colnames(Highest_VC_Biomarkers) = col_names_VT


# 3  - Depicting the drug responses of these genes with a heatmap
# 3.1- Manipulate colnames to show only cancer type
Metadata                        = as.data.frame(Metadata)
Metadata_Vorinostat             = Metadata    [ which(grepl  ("vorinostat" , Metadata$drug) ), ] 
Metadata_Vorinostat             = Metadata_Vorinostat [ order(Metadata_Vorinostat$cell), ,drop = FALSE ] 
Metadata_V_after                = subset(Metadata_Vorinostat, Metadata_Vorinostat$dose == "5000nM")
Metadata_V_after                = Metadata_V_after [, -which(colnames(Metadata_V_after) %in% 
                                                 c( "time", "drug", "sample", "dose" ))]
rownames(Metadata_V_after)      = Metadata_V_after$cell

colnames(Highest_VC_Biomarkers) = Metadata_V_after$tissue
```

```{r,fig.height=5, fig.width= 10}
pheatmap(Highest_VC_Biomarkers, main = "Expression Changes of Genes in 59 Cell Lines")
```

<div style="text-align: justify">
The heatmap shows that the idea of finding biomarkers which show cancer type specific responses did not work as expected. Only some genes, e.g. CD24 and AKR1C3 in melanoma samples, showed clearly cancer-specific responses.
</div>

## 3.3. Biomarkers received from FC values {#anchor7}

### 3.3.1.General biomarkers {#anchor8}
<div style="text-align: justify">
The FC value is a measure of the change in gene expression. Therefore, it is also well suited for the determination of biomarkers. In order to avoid mutually canceling effects and to find the largest change (regardless of its direction), we only use absolute values here.
</div>
```{r}
FC <- TreatedVorinostat - UntreatedVorinostat

FCVorinostatabs= abs(FC)
FCVorinostatmean <- apply(FCVorinostatabs, 1, mean)

#sort the values to get the 100 largest values:
sortedgeneralbiomarker <- sort(FCVorinostatmean, decreasing = TRUE)
sortedgeneralbiomarker <- as.matrix(sortedgeneralbiomarker)

#select the top 100 general biomarkers:
top100generalbiomarkers = sortedgeneralbiomarker[1:100,]
top100generalbiomarkers <- as.matrix(top100generalbiomarkers)
head(top100generalbiomarkers)
```

Create a vector with gene names:
```{r}
generalbiomarkergenes = row.names(top100generalbiomarkers)
```
<div style="text-align: justify">
Now we have determined biomarkers via the fold change. However, since we have worked only with absolute values, we cannot say whether our genes are up- or downregulated.
</div>

### 3.3.2. Up- and Downregulation of biomarkers {#anchor9}
<div style="text-align: justify">
We determine rowmeans of FC values. But unlike before, we do not use absolute values to calculate the averages. Consequently, we get a general overview which genes are mostly up- or downregulated.
</div>
```{r}
FC_meanrow= rowMeans(FC)

# look at the absolute values:
FC_meanrow = abs(FC_meanrow)

# sort the FC 
sortedFC <- sort(FC_meanrow, decreasing = TRUE)
sortedFC <- as.matrix(sortedFC)

# take the first 100 values for comparison 
FC100<- sortedFC[1:100,]
FC100 <- as.matrix(FC100)

# create a vector with genenames 
biomarkergenes_noabs = row.names(FC100)
```
<div style="text-align: justify">
We want to compare the top biomarkers with and without using absolute values to determine how much they differ:
</div>
```{r}
setequal(biomarkergenes_noabs,generalbiomarkergenes)
```

How many of the biomarkers are different? 

```{r}
# biomarkers in biomarkergenes_noabs but not in generalbiomarkergenes
diff= setdiff(biomarkergenes_noabs,generalbiomarkergenes)
length(diff)
diff
```
<div style="text-align: justify">
We see that only 5 genes differ in the top biomarkers. Nevertheless, it is better to trust the absolute values, as they indicate the largest change in gene expression. So we avoid, that up- or down-regulating effects cancel each other out.

In order to be able to work better with the biomarkers, we creat a matrix with the information, whether the genes are up or down regulated.  
</div>
```{r}
FCVorinostatmean_noabs= rowMeans(FC)

#Matrix with information about general up- or downregulation:
FCVorinostatmean_noabs_matrix <- as.matrix(FCVorinostatmean_noabs)

#loop, which adds those biomarkergenes to a vector, which are mostly upregulated:
i=1
Upregulated <- c()
while(i<=100) {
  if(FCVorinostatmean_noabs_matrix[generalbiomarkergenes[i],1] >0) {
    Upregulated <- union(Upregulated, generalbiomarkergenes[i])
  }
  i=i+1}

#Create a vector that includes for each gene sequentially if it is up- or downregulated:
i=1
Generalchange <- c()
while(i<=100) {
  if(FCVorinostatmean_noabs_matrix[generalbiomarkergenes[i],1] >0) {
    Generalchange <- append(Generalchange, "Up")
  } else {
    Generalchange <- append(Generalchange, "Down")
  }
  i=i+1}
```

Bind the information about up/downregulation as a new column to the biomarkermatrix:
```{r}
top100generalbiomarkers_withUporDown <- cbind(top100generalbiomarkers, Generalchange)
# rename
colnames(top100generalbiomarkers_withUporDown)[1] <- "FCmean"
head(top100generalbiomarkers_withUporDown)
```




### 3.4. Comparison of the differently received biomarkers {#anchor11}
<div style="text-align: justify">
Now we have determined biomarkers via the 2-sided T-test and want to investigate their FC values, since our biomarkers should cause the largest change in gene expression.

Do we have exactly the same biomarkers?
</div>
```{r}
setequal(generalbiomarkergenes,biomarkers.genes)
```

How many of the biomarkers are different?
```{r}
diff= setdiff(generalbiomarkergenes,biomarkers.genes)
length(diff)
```
<div style="text-align: justify">
Obviously, the biomarkers almost do not match at all. Thus, we want to compare p.Values and FC values in a new matrix which contains both values:
</div>
``` {r include = FALSE}
library(knitr)
```

```{r,results='hide',message=FALSE}
#sort FC matrix according biomarkers from pValues 
FC=TreatedVorinostat-UntreatedVorinostat
FC_m=rowMeans(FC)
FC_m=as.data.frame(FC_m)
FC =FC_m[biomarkers.genes,]
FC=as.data.frame(FC)
```

```{r}
pV_FC= cbind(FC,biomarkers.pValues)
pV_FC=as.data.frame(pV_FC)
head(pV_FC)
```
<div style="text-align: justify">
Now we know that p.Value and FC value do not deliver the same biomarkers. In fact, 87 out of 100 biomarkers are different.
If we compare the values directly, we see that a low p.value does not correlate with a high FC value, if we compare it with that of the biomarkers (see 3.3.2). This is because small values are also assigned small p.values by the T-test and the result of the test is therefore falsified. We have also seen before that almost all genes were found to be significant in the T test, but this is not credible. 
For this reasons, we cannot trust the biomarkers from the two-sided T-test. We will trust the biomarkers we received via the __fold change__ values, because they give us the genes that have been most altered.
</div>

***


## 4. Visualizations of biomarkers {#anchor12}

### 4.1. Density plot of biomarker gene expression {#anchor13}
<div style="text-align: justify">
Now we can repeat the the __gene-expression density plot__ from the beginning of our specific analysis with our biomarkers. We can see, that in terms of our identified genes Vorinostat leads to a clearly altered expression:
</div>
```{r,fig.height= 5, fig.width= 7}

Treated_Vorinostat_biomarker <- Treated[generalbiomarkergenes,TreatedVorinostatcolumns]
Untreated_Vorinostat_biomarker <- Untreated[generalbiomarkergenes,UntreatedVorinostatcolumns]

par(mar=c(5, 4, 5, 9))
plot(density(Untreated_Vorinostat_biomarker),col="blue" ,xlab = "Gene expression",
     main = "Effects of Vorinostat on biomarker gene expression")

lines(density(Treated_Vorinostat_biomarker), col = "red")

legend("topright", inset = c(-0.4,0), legend=c("Untreated", "Treated") , xpd = TRUE, 
       pch=19, col = c("blue", "red"))
```


### 4.2. Barplots of up- and downregulation of biomarkers {#anchor14}

```{r,fig.height= 5, fig.width= 7}
top100generalbiomarkers_withUporDown=as.data.frame(top100generalbiomarkers_withUporDown)
change=top100generalbiomarkers_withUporDown$Generalchange
col=palette(rainbow(2))
barplot(table(change), ylab="counts", main="Number of up and down regulated genes", col=col)
```
<div style="text-align: justify">
We see that we have approximately the same number of up and down regulated genes.


Another option to visualize the up- and downregulation of the biomarkers:
</div>
```{r plot_data, fig.height= 5, fig.width= 7}
FC=TreatedVorinostat-UntreatedVorinostat
FC_meanrow= rowMeans(FC)
FC_abs= abs(FC_meanrow)

FC_both= cbind(FC_meanrow,FC_abs)
FC_both=as.data.frame(FC_both)

# order this matrix 
FC_both_sorted <- FC_both[order(FC_both$FC_abs, decreasing = TRUE),]

# FC values of biomarkers
# take the first 100 of the sorted matrix, should be the same as un biomarkers_FC 
biomarkers_FC_values = FC_both_sorted[1:100,]
# remove the absolute values
biomarkers_FC_values <- subset( biomarkers_FC_values, select = -FC_abs)
biomarkers_FC_values = as.data.frame(biomarkers_FC_values)

par(mar=c(5, 4, 5, 9))
barplot( height= biomarkers_FC_values$FC_meanrow, 
         main= "Gene Expression Change of biomarkers",
         ylab="mean Fold change", xlab="biomarkers")

```

We see that most values are very similar to each other. 


### 4.3. Visualization of biomarkers with Vulcanoplot {#anchor15}

First we perform a one-sample t.test with normalized FC data:
```{r}
T.testFC <- apply(FC, 1, t.test)
```

Save the results in a data frame:
```{r}
TResults <- lapply(T.testFC, function(.tres) {      
  
  data.frame(t.value=.tres[1],dfs=.tres[2],conf.int1=.tres$conf.int[1],conf.int2=
               .tres$conf.int[2],p.value=.tres[3])
})

Tfinalresults <- do.call(rbind, TResults)
```

For the vulcano plot, a package needs to be installed: 
if (!requireNamespace('BiocManager', quietly = TRUE))
install.packages('BiocManager')
BiocManager::install('EnhancedVolcano')
```{r message=FALSE, warning=FALSE}
library(EnhancedVolcano)
```

Since we are interested in general biomarkers, we use the mean FC of the columns:
```{r}
log2FC <- apply(FC, 1, mean)
```

Add the FCvalues as a new column to the data frame containing the results of the t.test:
```{r}
FCandpvalues <- cbind(Tfinalresults, log2FC)
```
<div style="text-align: justify">
Create a Vulcanoplot:

First, all genes are assigned to the color black. Subsequently, those genes with a FC equal or larger than  the smallest FC of the top 100 FC-values are colored in gold. Those, which exceed the same limit in the negative range are assigned to the color green.
</div>
```{r, fig.height= 5, fig.width= 7}
keyvals <- rep('black', nrow(FCandpvalues))
names(keyvals) <- rep('FC<Top100', nrow(FCandpvalues))

keyvals[which(FCandpvalues$log2FC >= min(FC100))] <- 'gold'
names(keyvals)[which(FCandpvalues$log2FC >= min(FC100))] <- 'positive FC in Top100'

keyvals[which(FCandpvalues$log2FC <= -min(FC100))] <- 'green'
names(keyvals)[which(FCandpvalues$log2FC <= -min(FC100))] <- 'negative FC in Top100'

EnhancedVolcano(FCandpvalues,
                lab = rownames(FCandpvalues),
                x = 'log2FC',
                y = 'p.value',
                title = 'Significance versus fold change of all genes',
                selectLab = biomarkergenes_noabs,
                transcriptLabSize = 1.8,
                colOverride = keyvals,
                pCutoff = 10e-14,
                FCcutoff = min(FC100)
              )
```

Now we want to further investigate our biomarkers. 




## 5. Are the targets for vorinostat (HDACs) part of our biomarkers? {#anchor16}
<div style="text-align: justify">
As already mentioned in the introduction, the metadata describe the HDACs as targets of Vorinostat. Now we want to investigate whether these are among our biomarkers. To do this, we first create a vector that contains the target names. 
</div>
```{r}
###find target of vorinostat 

target=Drug_annotation$target
target=as.data.frame(target)

# take row with vorinostat 
target_vorinostat=target[9,]
target_vorinostat

```
<div style="text-align: justify">
We create a vector with targetnames as strings, because otherwise we have a problem with spacer |.  
</div>
```{r}
target_vorinostat=c("HDAC1","HDAC10","HDAC11","HDAC2","HDAC3","HDAC5","HDAC6",
                    "HDAC8","HDAC9")
```

Do we have HDACs in our biomarkers? 

```{r}
# search for HDAC in biomarkers 
target <- grep(pattern = "HDAC",colnames(top100generalbiomarkers))
length(target)
```
<div style="text-align: justify">
In drug annotation, HDAC genes are mentioned as target of vorinostat, but these genes are not in our defined biomarkers, so we want to check their FC values: 
</div>
```{r, include = FALSE}
library(dplyr)
```

```{r}
### find FC values of targets 

# create FC data 
FC_meanrow = as.data.frame(FC_meanrow)
genes = row.names(FC_meanrow)
FC_new=cbind(FC_meanrow,genes)


# filter data 
FC_target = as.data.frame(filter(FC_new, genes %in% target_vorinostat))

knitr::kable(head(FC_target), caption = "FC values vorinostat targets")
```



<div style="text-align: justify">
The changes in gene expression are really small. Most of the HDAC genes are up regulated insted of being down regulated. This could be because the cell wants to compensate the inhibition of the enzymes by their increased formation.
</div>

<div style="text-align: justify">
__Create boxplot of HDACs in comparison to biomarkers__
To compare the FC of the HDAC genes with some of our biomarkers, we create a vector which contains both biomarkers and HDAC-genes. 
</div>
```{r}
first10biomarkers= generalbiomarkergenes[1:10]
c=union(target_vorinostat,first10biomarkers)
```

Find the FC values of those genes in the vector in the FC matrix:
```{r}
FC= TreatedVorinostat-UntreatedVorinostat
FC=as.data.frame(FC)

FC_HDAC <-FC[c,]
```
<div style="text-align: justify">
Not all HDAC targets which were listed in the annotation data can be found in our expression data, so we remove the missing ones: 
</div>
```{r}
FC_HDAC =na.omit(FC_HDAC)
```

<div style="text-align: justify">
Now we plot a boxplot in order to visualize the much lower effect on HDAC-gene-expression compared to some of our biomarkers: 
</div>
```{r, fig.height= 5, fig.width= 7}
palette(rainbow(17))
boxplot(t(FC_HDAC), las=3, col = col, ylab="Fold Change Values", cex.axis= 0.65,  main="Comparison of Biomarker and target gene expression change ")
abline(v=7.5, col="blue", lty=5)
```
<div style="text-align: justify">
We can summarize, that the changes in HDAC gene expression are low. Some of the them are slightly upregulated. As already mentioned, it might be that treated cells try to compensate the inhibition of the enzymes by their increased production.
</div>


## 6. PCA with general-biomarker-matrix {#anchor17}

Select only those genes from the FC matrix which were identified as biomarkers:
```{r, fig.height= 5, fig.width= 7}
FCbiomarkers <- FC[generalbiomarkergenes,]
```

Execute the PCA:
```{r, fig.height= 5, fig.width= 7}
FCbiomarkers.pca = prcomp(FCbiomarkers, center=T, scale. = T)

#For coloring according to tissue type we use information from Metadata:
Metadata <- as.data.frame(Metadata)

#We only need the tissueinformation of those samples treated with Vorinostat.
#Since Metadata includes them twice, once with and once without drug treatment,
#we select only the Treated samplenames:
Vorinostatsamples <- grep(Metadata$sample, pattern= "vorinostat_5000nM")
Metadatavorinostattissue <- Metadata[Vorinostatsamples,"tissue"]

#Add the tissueinformation as a new row to the Biomarkermatrix:
FCbiomarkerswithtissue <- rbind(FCbiomarkers, Metadatavorinostattissue)

#save tissue information as factors so it can be used for coloring:
vorinostattissuefactor <- as.factor(FCbiomarkerswithtissue["Metadatavorinostattissue",])


#PC 3 & 4 group the tissues best but a clear grouping is not shown as well:
par(mar=c(5, 4, 5, 9))
palette(rainbow(9))
plot(FCbiomarkers.pca$rotation[, 3], FCbiomarkers.pca$rotation[, 4],
     col = Metadatavorinostattissue, pch = 19, xlab = "PC3", ylab = "PC4",
     main = "PCA with FC of biomarkers colored according to tissue")

#Add a legend to the plot:
levels <- as.factor(levels(vorinostattissuefactor))
legend("topright", inset = c(-0.3,0),levels(Metadatavorinostattissue),
       xpd = TRUE, pch=19, col = palette(rainbow(9)))
```


## 7. Find biomarkers only for leukemia-celllines {#anchor18}
<div style="text-align: justify">
Since we found in the beginning, that Vorinostat is used against T-cell lymphomas, we are interested in how Vorinostat influences especially the leukemia-celllines. Therefore, we now want to investigate these specific samples. 
</div>
```{r}
#We only need the tissueinformation of those samples treated with Vorinostat
#Since Metadata includes them twice, once with and once without drug treatment,
#we select only the Treated samplenames:
Vorinostatsamples <- grep(Metadata$sample, pattern= "vorinostat_5000nM")
Metadatavorinostattissue <- Metadata[Vorinostatsamples,"tissue"]

#Add the tissueinformation as a new row to the FCVorinostat-matrix:
FCwithtissue <- rbind(FC, Metadatavorinostattissue)

#select only those samples/celllines which belong to Leukemia:
FCwithtissue <- as.data.frame(FCwithtissue)
Leukemiasamples <- colnames(FCwithtissue[Metadatavorinostattissue== "Leukemia"])
FCLeukemia <- FCwithtissue[,Leukemiasamples]

#Now we do not need the tissue information any longer:
FCLeukemia <- FCLeukemia[-13300,]

#We want the FC-mean for each gene:
FCLeukemiaabs <- abs(FCLeukemia)
FCLeukemiameans <- apply(FCLeukemiaabs,1 ,mean)

#Sort the biomarkers that those with the biggest FC are on top:
sortedFCleukemiamean <- sort(FCLeukemiameans, decreasing = TRUE)
sortedFCleukemiamean <- as.matrix(sortedFCleukemiamean)

# take the first 100 as Leukemia-biomarkers:
biomarkersLeukemia = sortedFCleukemiamean[1:100,]
biomarkersLeukemia <- as.matrix(biomarkersLeukemia)

# create a vector with gene names of Leukemia-biomarkers: 
biomarkersLeukemiaGenes = row.names(biomarkersLeukemia)


#Comparison with general biomarkers:
intersect(biomarkersLeukemiaGenes, generalbiomarkergenes)

biomarkersLeukemiaGenes[biomarkersLeukemiaGenes == generalbiomarkergenes]
```


```{r,fig.height= 5, fig.width= 7}

Treated_Leukemiabiomarker <- Treated[biomarkersLeukemiaGenes ,TreatedVorinostatcolumns]
Untreated_Leukemiabiomarker <- Untreated[biomarkersLeukemiaGenes,UntreatedVorinostatcolumns]

par(mar=c(5, 4, 5, 9))
plot(density(Untreated_Leukemiabiomarker) ,xlab = "Gene expression", col = "blue",
     main = "Effects of Vorinostat on Leukemiabiomarker gene expression")
lines(density(Treated_Leukemiabiomarker), col = "red")
legend("topright", inset = c(-0.4,0), legend=c("Untreated", "Treated") , xpd = TRUE, 
       pch=19, col = c("blue", "red"))
```
<div style="text-align: justify">
The plot looks quite similar to the plot before which contained the general biomarkers. Since the biomarkers of leukemia are partly the same as the general biomarkers(38 out of 100 biomarkers), the similarity of the plots is not surprising.
</div>
