---
title: 'MAF: Visualization of somatic mutations'
author: "Florencia Zuniga"
date: "7/17/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<p> <div style="text-align: justify"> Which types of mutations affect the genes we explore in this analysis? Can we exploit this information to know which types of therapy might be effective? This analysis will focus on visualizing the somatic mutations of several genes and the samples they affect. We aim to find differences in patterns of mutations when we consider all genes and when we consider only 100 biomarkers. </div>

[1. LOADING DATA](#anchor1)

[2. Creation of polished MAF files ](#anchor2)

* [2.1 Creation of a polished MAF file for all genes](#anchor3)

* [2.2 Creation of a polished MAF file for 100 Biomarkers](#anchor4)

[2. General visualizations ](#anchor5)

* [2.1 Summaries of both MAF files)](#anchor6)

* [2.2 Oncoplots and oncostrips](#anchor7)

* [2.4 Somatic interactions](#anchor8)

[3. Drug-gene interactions: which types of therapy should be considered with the given mutations ](#anchor9)

* [3.1 Druggable categories)](#anchor10)

* [3.2 Pathways](#anchor11)

[4. General conclusions ](#anchor12)

[5. References ](#anchor13)




***




##  1.  LOADING DATA {#anchor1}                                                                                             
***
### Loading packages

```{r read_data, message = FALSE, warning = FALSE}
library(readr)
library(rstudioapi)
library(maftools)
```


```{r,include = FALSE}
wd = ("/GitHub/project-02-group-05")
```


Reading the data
```{r}

# Reading the data
Untreated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_untreated.rds"))
Treated <- readRDS(paste0(wd,"/data/NCI_TPW_gep_treated.rds"))

Metadata = read.table(paste0(wd,"/data/NCI_TPW_metadata.tsv"), 
                      header = TRUE, sep ="\t", stringsAsFactors = TRUE)

Sensitivity <- readRDS(paste0(wd,"/data/NegLogGI50.rds"))


Basal <- readRDS(paste0(wd,"/data/CCLE_basalexpression.rds"))
Copynumber <- readRDS(paste0(wd,"/data/CCLE_copynumber.rds"))
Mutations <- readRDS(paste0(wd,"/data/CCLE_mutations.rds"))

Metadata = read.table(paste0(wd,"/data/NCI_TPW_metadata.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)

Cellline_annotation = read.table(paste0(wd,"/data/cellline_annotation.tsv"), 
                                 header = TRUE, sep ="\t", stringsAsFactors = TRUE)
Drug_annotation = read.table(paste0(wd,"/data/drug_annotation.tsv"), 
                             header = TRUE, sep ="\t", stringsAsFactors = TRUE)

# Transforming the data

Treated <- as.data.frame(Treated)
Untreated <- as.data.frame(Untreated)
Sensitivity<- as.data.frame(Sensitivity)
```


Data normalization
```{r}
Untreated_norm <- apply(Untreated, 2, function(x){
  (x - mean(x)) / sd(x)
 })


Treated_norm <- apply(Treated, 2, function(x){
  (x - mean(x)) / sd(x)
 })


FC <- Treated - Untreated
FC_norm <- apply(FC, 2, function(x){
  (x - mean(x)) / sd(x)
 })
```


```{r, include=FALSE}
### Loading data of the biomarkers                                                                                                    
##  (1)  Creating Vorinostat

#Untreated matrix
UntreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Untreated))

#Same with treated matrix
TreatedVorinostatcolumns <- grep(pattern = "vorinostat",colnames(Treated))

#Define Vorinostat-data: 
UntreatedVorinostat <- Untreated[,UntreatedVorinostatcolumns]
TreatedVorinostat <- Treated[,TreatedVorinostatcolumns]

#fold change matrix 
FC <- TreatedVorinostat - UntreatedVorinostat

#Sensitivity 
vorinostat_Sensitivity_alleZeilen= grep ('vorinostat', rownames(Sensitivity))
vorinostat_Sensitivity_data= Sensitivity[vorinostat_Sensitivity_alleZeilen,]

##  (2)  Creating FC Data -  Finding the Biomarkers

FC <- TreatedVorinostat - UntreatedVorinostat

#We work with mean of the rows because we only want to compare the genes 
FC_meanrow= rowMeans(FC)

## Sorting the data 
#We work with absolute value to find the highest values, 
#because we want to have the most up and down regulated genes.
FC_abs= abs(FC_meanrow)

#We sort the values to get the 100 largest values 
sortedFC_abs <- sort(FC_abs, decreasing = TRUE)
sortedFC_abs <- as.matrix(sortedFC_abs)

#We select the first n for biomarkers 
biomarkers_FC30 = sortedFC_abs[1:30,]
biomarkers_FC30 <- as.matrix(biomarkers_FC30)

biomarkers_FC100 = sortedFC_abs[1:100,]
biomarkers_FC100 <- as.matrix(biomarkers_FC100)

## Creating a matrix with FC values, that are both positive and negative 
FC_both= cbind(FC_meanrow,FC_abs)
FC_both=as.data.frame(FC_both)

#Ordering this matrix 
FC_both_sorted <- FC_both[order(FC_both$FC_abs, decreasing = TRUE),]

#FC values of biomarkers: We select the first 100 of the sorted matrix. 
biomarkers_FC_values30 = FC_both_sorted[1:30,]

biomarkers_FC_values100 = FC_both_sorted[1:100,]

#Removing the absolute values
biomarkers_FC_values30 <- subset( biomarkers_FC_values30, select = -FC_abs)
biomarkers_FC_values30 = as.matrix(biomarkers_FC_values30)

biomarkers_FC_values100 <- subset( biomarkers_FC_values100, select = -FC_abs)
biomarkers_FC_values100 = as.matrix(biomarkers_FC_values100)

```

***

##  2. Creation of polished MAF files {#anchor2}

<p> <div style="text-align: justify"> For the 'CCLE_mutations' file to be read using the 'maftools' package, we must first edit it so that the resultung 'table' has all the mandatory columns. </div>

```{r}
Mutations <- readRDS(paste0(wd,"/data/CCLE_mutations.rds"))

names(Mutations)[names(Mutations) == "Tumor_Seq_Allele1"] <- "Tumor_Seq_Allele2"
names(Mutations)[names(Mutations) == "Start_position"] <- "Start_Position"
names(Mutations)[names(Mutations) == "End_position"] <- "End_Position"
rownames(Mutations) <- c()
MutationsT <- Mutations[,c(1,4,5,6,10,11,8,9,16,2,3,7,12,13,14,15,17,18)]
```


### 2.1 Creation of a polished MAF file for all genes {#anchor3}

```{r echo = T, results = 'hide'}
write.table(MutationsT, file = "MutationsT.csv", row.names = F, sep = "\t")

laml <- read.maf(maf ="C:/GitHub/project-02-group-05/MutationsT.csv", useAll = T, verbose = T)
```


### 2.2 Creation of a polished MAF file for 100 biomarkers {#anchor4}

<div style="text-align: justify"> To create this table, the biomarkers found using the fold change in the first part of the specific analysis are used. </div>

```{r echo = T, results = 'hide'}
BM_mut = MutationsT[ which((MutationsT$Hugo_Symbol) 
                                  %in% rownames(biomarkers_FC_values100)), ]

write.table(BM_mut, file = "BM_mut.csv", row.names = F, sep = "\t")

BM_laml <- read.maf(maf ="C:/GitHub/project-02-group-05/BM_mut.csv", useAll = T, verbose = T)
```
***

## 2. General visualizations {#anchor5}
***
### 2.1 Summaries of both MAF files {#anchor6}
<p> 
<p> <div style="text-align: justify">  To begin our analysis, we visualize different plots that summarize the most important informations. </div>
<p> 
**Summary for all genes**
```{r, fig.align= 'center'}
plotmafSummary(maf = laml, 
               rmOutlier = TRUE, 
               addStat = 'mean', 
               dashboard = TRUE, 
               titvRaw = FALSE)

```

**Summary for the biomarker genes**

<p> <div style="text-align: justify"> As the code for some plots only differs in the MAF file that is to be read or other small details, only the plots and not the code is shown in those cases to avoid redundancy. </div>

```{r, echo=FALSE, fig.align= 'center'}
plotmafSummary(maf = BM_laml, 
               rmOutlier = TRUE, 
               addStat = 'mean', 
               dashboard = TRUE, 
               titvRaw = FALSE)
```

<p> <div style="text-align: justify"> We can already observe that plenty of differences arise when we consider all genes vs 100 biomarkers. Although the most common type of mutation is the missense one in both cases, in the summary of all genes more types of mutations affect the genes in the samples. Only 4 distinct type of mutations can be observed for the biomarkers. </div>


<p> <div style="text-align: justify"> Another interesting observation, is that when all genes are considered, there are mutations that happen because of an insertion. This is no longer true when we only consider the biomarkers. However, in both cases the variant type plot shows that the most common type of variation is the single nucleotide polymorphism. </div>

<p> <div style="text-align: justify"> If we look at the plot 'Variants per sample' in both summaries, another surprising result is that there is a drastic change in the mean. This is probably a result of less variation of variation when we consider only the biomarkers, as we would hope those genes would act in a similar way, because the biomarkers are the most affected genes by vorinostat in general in all samples. </div>

<p> <div style="text-align: justify"> The top 10 mutated genes are all different in the two summaries. It is worth nothing that 2 of the top ten genes affected when only considering the biomarkers are related to the expression of histones. These genes are **HIST1H1C**, which encodes the protein Histone H1.2, and the gene **HIST1H2BD**, which encodes the protein Histone H2B type 1-D. This finding is of particular relevance, because the mechanisms of vorinostat is to inhibit HDACs. </div>
***
### 2.2 Oncoplots and oncostrips {#anchor7}

<p> <div style="text-align: justify"> We can also plot the top n genes that shows the most variation in the samples. For this, we use waterfall plots that show the different mutations that affect each sample on the top, the percentage of samples that have been affected by mutations in certain genes on the right, the names of the mutated genes on the left and the transitions/transversions of nucleotides per sample on the bottom. </div>

#### Ploting oncoplot with Transversions/Transitions

**All genes**

```{r, fig.align= 'center'}
oncoplot(maf = laml, top = 15, draw_titv = TRUE)

```

<p> <div style="text-align: justify"> As all the samples we are handling are of cancerous nature, all of them have mutations in one or several genes, with the most common mutation being a missense one and the most common transition of nucleotide being the one from cytosine to thymine. </div>
<p>
<p> **Biomarker genes**
<p>
```{r, echo=FALSE, fig.align= 'center'}
oncoplot(maf = BM_laml, top = 15, draw_titv = TRUE)

```

<p> <div style="text-align: justify">  When we consider only the biomarkers, within the top 15 genes there are only 21 samples that are mutated. What is also interesting, is that when we consider 100 biomarkers only 28 samples of the 58 that were studied in total show some type of mutation. There is less variation in the type of mutations than in the previous oncoplot. Most biomarkers have one or several missense mutations, with other kinds of mutations appearing only rarely. </div>
<p>
***
#### Oncostrips
<p>
<p> <div style="text-align: justify"> Are there mutational differences between the top 10 biomarkers and the 10 genes with the least change in expression? These genes are obtained from the matrixes in the first part of the specific analysis when the biomarkers were found.</div>
<p>
##### **Oncostrip of the top 10 biomarkers**

```{r, fig.align= 'center'}
oncostrip(maf = laml, 
          genes = c('DHRS2',
                    'ABAT', 
                    'SERPINI1', 
                    'MIR612///NEAT1	',
                    'HBA2///HBA1	',
                    'CLU',
                    'NMI',
                    'STC1',
                    'AREG',
                    'NSMAF',
                    'SERPINH1'),
          showTumorSampleBarcodes = TRUE,
          SampleNamefontSize = 0.8
          )
```


##### **Oncostrip for the 10 genes with the least change in expression**


```{r, echo=FALSE, fig.align= 'center'}
oncostrip(maf = laml, 
          genes = c('ADAMTS6', 
                    'YWHAQ',
                    'EMID1', 
                    'PLCH2', 
                    'HRH2',
                    'WNT10B',
                    'RGPD6///RGPD8///RGPD3///RGPD4///RGPD5',
                    'ANGPTL8',
                    'CABP2',
                    'UPK2',
                    'CLPS'),
          showTumorSampleBarcodes = TRUE,
          SampleNamefontSize = 0.8
        )
```

<p> <div style="text-align: justify"> The results we observe in these plots coincide with our previous findings: when we consider the biomarkers, in this case the top 10, most of them are affected by missense mutations, which are point mutations that causes the formation of a codon that codes for a different aminoacid. The second oncostrip shows more variation in the types of mutation present. </div>
<p>
***
### 2.3 Somatic interactions {#anchor8}

<p> <div style="text-align: justify"> In this portion of the analysis we explore the somatic interactions of genes, in that we observe which genes are co-occuring and wich ones show strong exclussiveness. To find such genes, the function relies on pairwise Fisher’s Exact tests. </div>
<p>
**All genes**
<p>
#### Somatic Interactions
```{r, include=TRUE, results='hide', fig.align= 'center'}
somaticInteractions(maf = laml, top = 30, pvalue = c(0.05, 0.1))
```
<p> <div style="text-align: justify"> Here we see that although co-occurance is a rather common trait, exclusivity is rare. When we consider 30 genes, exclusivity happens only 3 times and it always includes the gene **MT-ND5**, which encodes for the protein NADH-ubiquinone oxidoreductase chain 5. This gene is one of the top 10 mutated genes, which can be seen in the summary for all genes at the begining of this analysis. The gene is altered in 41% of the samples. </div>
<p>

#### Oncostrip
<p> <div style="text-align: justify"> Here we plot 2 genes which are mutually exclusive (HERC2 and MT-ND5) and two genes that are co-occurrent (USH2A and XIRP2). </div>

<p> The function of each gene can be obtained in the table below: 
<p> 
|   | Codes for   | Protein name  | Function  |   
|:-:|:-:|:-:|:-:|
| HERC2 | Protein  |  E3 ubiquitin-protein ligase  | DNA repair regulation |   
| MT-ND5 |Protein | NADH-ubiquinone oxidoreductase chain 5  | Part of the respiratory complex I|   
| USH2A  | Protein | Usherin | Component of basement membranes  |   
| XIRP2  |  Protein | Xin actin-binding repeat-containing protein 2 | Protects actin filaments from depolymerization |  
|   |   |   |   |
<p> 



```{r, fig.align= 'center'}
oncostrip(maf = laml, genes = c('USH2A', 'XIRP2','MT-ND5', 'HERC2'))

```
<p> <div style="text-align: justify"> Both pairs show different patterns of mutations. Some mutations can be observed in each gene of the pair in the same sample. </div>
<p>
<p>

**Biomarker genes**
<p>
#### Somatic Interactions

```{r, include=TRUE, results='hide', fig.align= 'center'}
somaticInteractions(maf = BM_laml, top = 30, pvalue = c(0.05, 0.1))

```
<p> <div style="text-align: justify">  Now that only the biomarkers are being considered, we see that for the 30 shown genes co-occurrence is rare and exclusivity does not appear. As less than 30 genes have co-occurrence, less than 30 genes are shown, eventhough the code asks for 30. Instead 18 genes appear in the plot. </div>
<p>

#### Oncostrip
<p> <div style="text-align: justify"> As there is only strong co-occurrence between two genes (NSMAF and DHRS2), we plot these two in the oncostrip along with the LMNB1 gene that has no co-occurence and no exclusivity . </div>

<p> The function of each gene can be obtained in the table below: 
<p> 
|   | Codes for   | Protein name  | Function  |   
|:-:|:-:|:-:|:-:|
| NSMAF | Protein  | FAN  | Neutral sphingomyelinase activation |   
| DHRS2 |Protein | Dehydrogenase/reductase SDR family member 2  | Metabolism of different compunds|   
| LMNB1  | Protein | Lamin B1 | Intermediate filament protein  |   
|   |   |   |   |
<p> 


```{r, echo=FALSE, fig.align= 'center'}
oncostrip(maf = BM_laml, genes = c('NSMAF', 'DHRS2', 'LMNB1'))

```
<p> <div style="text-align: justify">The first two genes only show the same type of mutation in ine cell line. </div>

<p> <div style="text-align: justify">As very few events of co-occurence are observed when we consider only the biomarkers, we can conclude that there are few genes that affect various samples when another gene is affected by a mutation. This, of courses, poses the question as to how biomarkers and diferent cell lines relate to each other and among themselves, and of whether we would see clusters if we plotted both in a heatmap. </div>


***
## 3. Drug-gene interactions: which types of therapy should be considered with the given mutations {#anchor10}
***
<p> <div style="text-align: justify"> The [The Drug Gene Interaction Database](http://www.dgidb.org/) defines drug-gene interaction as a known interaction of a gene and a drug. One category here is 'druggable gene. This is a set of genes that are believed to be most likely be treatble with certain drugs. Genes in these categories may or may not have existing drugs that target do them. </div>
***

### 3.1 Druggable categories {#anchor9}

**All genes**

```{r, fig.align= 'center'}
dgi = drugInteractions(maf = laml, fontSize = 0.75)

```
<p> <div style="text-align: justify"> The plot shows druggable categories and up to 5 genes that might be involved in the gene category. 

<p>  <div style="text-align: justify"> One category called  'the druggable genome' represens the subset of about 30,000 genes in the human genome that express proteins that are capable of binding to drug-like molecules. </div>

**Biomarker genes**

```{r, echo= FALSE, fig.align= 'center'}
dgi = drugInteractions(maf = BM_laml, fontSize = 0.75)

```

<p> <div style="text-align: justify"> We can also explore other details about the druggable genes, such as the existing drugs and the mechanism of action. </div>

```{r}
dnmt3a.dgi = drugInteractions(genes = "AREG", drugs = TRUE)
dnmt3a.dgi[,.(Gene, interaction_types, drug_name, drug_claim_name)]
```

<p> <div style="text-align: justify"> Surprisingly, there are more druggable genes related to histone modification when we consider all genes tha when we consider only the biomarkers. This is suprising because of the mechanism of action of vorinostat. Nonetheless, more genes of the category 'clinically actionable' related to histone expression appear when we only consider the biomarkers. Therefore our results still coincide with the the information in the drug annotation about our drug of interest. </div>


***
### 3.2 Pathways {#anchor11}

<p> <div style="text-align: justify"> In this last part of the analysis we focus on the oncogenic pathways that result from the mutations. As we see in the table below there are 10 possible pathway categories ("Cell Cycle","Hippo", "MYC", "NOTCH", "NRF2", "PI3K"       "RTK-RAS"    "TGF-Beta", "TP53", and  "WNT"). Number of occurences  and of affected genes appear aswell. </div>

**All genes**


<p>
```{r}
OncogenicPathways(maf = laml)

```
<p>
<p> <div style="text-align: justify"> We can also plot the genes affected by the pathway and the samples in which these genes are mutated. </div>
<p>
```{r}
PlotOncogenicPathways(maf = laml, pathways = "PI3K")

```
<p>
<p> <div style="text-align: justify"> Tumor suppressor genes are shown in red and oncogenes and shown in blue. Mutations in the PI3K pathway are common for our data, appearing once or several times per sample. </div>
<p>
**Biomarker genes**
<p>

```{r, echo=FALSE}
OncogenicPathways(maf = BM_laml)

```

<p> <div style="text-align: justify">  None of these pathways are affected in the biomarkers genes. </div>

***

## 4. General conclusions {#anchor12}
***

<p> <div style="text-align: justify">  There are clear differences between the mutations of all genes and the mutations of the biomarkers. The most obvious one is in the level of mutational variation for both categories, with the level being much higher for all genes. For the biomarkers, we observe amost exlusively missense mutations. Another interesting finding of this exploration is the relevance of genes related to histones, that appear in relevant positions several times during this analysis. </div>



***
## 5. References {#anchor13}
***
<p> <div style="text-align: justify"> Gene Cards. 2019. GeneCards®: The Human Gene Database. [ONLINE] Available at: https://www.genecards.org/. [Accessed 17 July 2019]. </div>


<p> <div style="text-align: justify"> National Human Genome Research Institute. 2019. Missense Mutation. [ONLINE] Available at: https://www.genome.gov/genetics-glossary/Missense-Mutation. [Accessed 17 July 2019]. </div>


<p> <div style="text-align: justify"> The Drug Gene Interaction Database. 2019. FAQ. [ONLINE] Available at: http://www.dgidb.org/faq. [Accessed 17 July 2019].  </div>


<p> <div style="text-align: justify"> Mayakonda A, Lin DC, Assenov Y, Plass C, Koeffler HP. 2018. Maftools: efficient and comprehensive analysis of somatic variants in cancer. Genome Resarch. </div>


<p> <div style="text-align: justify">  Hopkins, A., & Groom, C. (2002). The druggable genome. Nature Reviews Drug Discovery, 1(9), 727-730. doi: 10.1038/nrd892 </div>





